---
title: "Vietnam_CRCmicrobiome_analysis_2021.Rmd"
author: "Hao Chung The"
date: "10/12/2021"
output: html_document
---

This dataset is composed of saliva and gut mucosal 16S profiling of 25 controls and 43 CRC cases (totaling 203 samples). DNA from each sample is extracted and subjected to 16S sequencing using the conventional V4 primers. The amplified region is ~250 bp in length, and the data has been denoised and merged to create ASVs using DADA2 (in a separate R file). The amplicon sequence variant (ASV) approach is chosen instead of clustering sequences into OTUs using fixed threshold. 

Taxonomic assignment was performed under the RDP implementation in DADA2, using the SILVA training dataset, with 50% bootstrap confidence threshold. ASVs were classified up to the species level where possible. 
---

=========================================================
We first import the phyloseq object after pre-processing and associated metadata
Initialize by loading necessary libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(phyloseq)
library(stringr)
library(vegan)
library(phytools)
library(ranacapa)
library(ggplot2)
library(philr)
library(zCompositions)
library(DESeq2)
library(microbiome)
library(ANCOMBC)
library(dplyr)
library(nloptr)
library(randomForestSRC)
library(ggRandomForests)
library(reshape2)
library(zinbwave)
library(scran)
library(apeglm)
library(ashr)
library(corncob)
library(qdap)
library(cluster)
library(fpc)
library(SpiecEasi)
library(igraph)

load(file="./27EN_phyloseq_CRC_keep.RData")
meta.df <- read.table(file="./27EN_compiled_metadata.csv", sep=",", header=TRUE, stringsAsFactors = FALSE)
rownames(meta.df) <- meta.df$sample_ID
meta.df <- meta.df[,-c(1)]

## attach metadata into the phyloseq object 
sample_data(crc.keep) <- sample_data(meta.df)
crc.keep ## 2461 taxa with 203 samples

## Plotting simple PCA for sanity check. The dataset includes both saliva and gut mucosal microbiome data
sample_data(crc.keep)$sample_type <- 
      factor(sample_data(crc.keep)$sample_type, levels=c('saliva', 'biopsy', 'polyp', 'nontumour', 'tumour'))
all.ord <- ordinate(crc.keep, method="PCoA", distance='bray') ## Bray-curtis distance
plot_scree(all.ord) + xlim(as.character(seq(1,50))) + coord_cartesian(xlim=c(0,50))
all.ord$values$Relative_eig[1:2] ## the two principal coordinates account for 22.5% and 6.9% of variation in the data

plot_ordination(crc.keep, all.ord, type="samples", color="sample_type") + geom_point(size=2) + 
      theme_gray() + scale_color_manual(values=c("forestgreen","lightskyblue",'royalblue3', "lightpink1", "indianred4")) + 
      xlab("PCoA1 (22.6%)") + ylab("PCoA2 (6.9%)") + 
      theme(legend.position = "bottom",legend.background = element_rect(color='gray90')) 
### This showed that saliva samples are well separated from gut mucosa samples since they are composed of completely different microbiomes. This showed that there has been no gross mislabelling during the sequencing

```

Compare the baseline characteristic between the two groups

```{r}
## first check the data
with(meta.df, table(sample_type))
patient.df <- meta.df[meta.df$sample_type == 'tumour' | meta.df$sample_type=='polyp',]
patient.df[is.na(patient.df$Age),] ## five patient with missing data. These will be removed.
na.patient <- patient.df[is.na(patient.df$Age),'Patient']
patient.df <- patient.df[!(patient.df$Patient %in% na.patient),]
table(patient.df$Group) ## 21 control and 42 cases with associated metadata

## Testing for multiple demographic and clinical information
with(patient.df, t.test(Age~Group)) ## NS
fisher.test(with(patient.df, table(sex, Group))) # NS
with(patient.df, t.test(BMI~Group)) #NS
fisher.test(with(patient.df, table(Asian_BMI_class, Group))[c(1:2),]) #NS
fisher.test(with(patient.df, table(diabetes, Group))) #NS
fisher.test(with(patient.df, table(HBP, Group))) #NS HBP= high blood pressure
fisher.test(with(patient.df, table(smoking_active_2y, Group))) #NS
fisher.test(with(patient.df, table(alcohol_active_2y, Group))) ## NS

## Any self-reported oral illnesses
patient.df$oral_dis <- as.vector(patient.df$gingivitis == 'yes' | patient.df$periodontitis == "yes" | 
                                                    patient.df$badbreath=='yes')
patient.df$oral_dis <- str_replace(patient.df$oral_dis, 'TRUE', 'yes')
patient.df$oral_dis <- str_replace(patient.df$oral_dis, 'FALSE', 'no')
fisher.test(with(patient.df, table(oral_dis, Group))) # NS

# Family history of cancer
patient.df$hist_cancer <- patient.df$family_history_cancer
patient.df[which(patient.df$hist_cancer != 'no'), 'hist_cancer'] <- 'yes'
fisher.test(with(patient.df, table(hist_cancer, Group))) ## NS 

# Regarding anatomy
with(patient.df, table(anatomy, Group))

## There is no signficant differences in the metadata between cases and controls
```

We will analyze the saliva and gut mucosal microbiomes separately

```{r}
## First we output the saliva microbiome first
saliva.ps <- prune_samples(sample_names(crc.keep)[which(sample_data(crc.keep)$sample_type == 'saliva')], crc.keep)
saliva.ps <- prune_taxa(taxa_sums(saliva.ps) > 0, saliva.ps)
saliva.ps ## 67 samples with 1278 taxa

## explore singleton
saliva.tx.pres <- rowSums(abundances(saliva.ps) > 0) ## taxa with number of non-zero presences in the samples.
saliva.singleton <- as.vector(as.matrix(tax_table(saliva.ps)[names(which(saliva.tx.pres == 1)),"Genus"])) ## 554 singleton taxa
names(saliva.singleton) <- rownames(tax_table(saliva.ps)[names(which(saliva.tx.pres == 1)), "Genus"])
sort(table(saliva.singleton),decreasing = TRUE) ## most common singleton taxa are Prevotella, Leptotrichia, Veillonella, Fusobacterium, Treponema, etc. 

## Examine the counts of singleton taxa 
sort(taxa_sums(saliva.ps)[names(saliva.singleton)], decreasing=TRUE)
summary(taxa_sums(saliva.ps)[names(saliva.singleton)]) # median is 38 reads, third quartile is 79 reads.
## Only retain singleton with high counts, select ones with read count ≥ 79 reads
table(saliva.singleton[names(which(taxa_sums(saliva.ps)[names(saliva.singleton)] < 79))]) ## a wide range of genera
saliva.taxa.rm <- names(which(taxa_sums(saliva.ps)[names(saliva.singleton)] < 79))

## Remove singleton taxa that have read abundance < 79 reads
saliva.fil <- prune_taxa(!(taxa_names(saliva.ps) %in% saliva.taxa.rm), saliva.ps)
saliva.fil ## 865 taxa
summary(sample_sums(saliva.fil)/sample_sums(saliva.ps)) ## only 1-2% difference in the total number of reads, median of 99.5% retainment 

## Check if the samples have necessary coverage/library size, by using rarefaction curve
saliva.ggrare <- ggrare(saliva.fil, step=100, label=NULL, color='Group')
saliva.ggrare <- saliva.ggrare + xlim(0,10000) + theme(legend.position = 'bottom')
saliva.ggrare ## plateauing occurs at less than 2,500 reads
sort(sample_sums(saliva.fil)) ## sample S0-5 only have 837 reads, to remove for downstream analyses

saliva.fil <- prune_samples(sample_names(saliva.fil) != 'S0-5', saliva.fil)
saliva.fil <- prune_taxa(taxa_sums(saliva.fil) > 0, saliva.fil)
```

For downstream analyses, we input a phylogenetic of 865 taxa found in the saliva dataset. The phylogeny was constructed by IQtree (GTR+G model, 1000 rapid bootstrap, midpoint rooted), using the PASTA-generated alignment of the bacterial V4 sequences of 865 ASVs. 

Some exploratory analyses for saliva microbiome, using basic ordination.

```{r}
saliva.tre <- read.newick(file="./27EN_saliva_fil_865seqs.snp.fa.contree.midroot.newick")
phy_tree(saliva.fil) <- saliva.tre
saliva.fil
sample_data(saliva.fil)$oral_dis <- patient.df[match(sample_data(saliva.fil)$Patient, patient.df$Patient), 'oral_dis']
## Remove samples without associated metadata
saliva.fil2 <- prune_samples(!(is.na(sample_data(saliva.fil)$Age)), saliva.fil)
saliva.fil2 <- prune_taxa(taxa_sums(saliva.fil2) > 0, saliva.fil2)
saliva.fil2  ## 62 samples 

## Transform phyloseq into normalized count based on the median of sample_sums
saliva.median <- round(median(sample_sums(saliva.fil2))) #36250 reads
standf <- function(x, t=saliva.median) round(t* (x/sum(x)))
saliva.prop <- transform_sample_counts(saliva.fil2, standf) ## normalized to the same library size
saliva.dt <- transform_sample_counts(saliva.fil2, function(x) x/sum(x)) ## relative abundance

## Examining the composition of saliva microbiome
plot_bar(saliva.prop, fill="Phylum") + geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="stack")

plot_richness(saliva.prop, measures=c("Shannon", "Chao1"), color="Group") ## quite similar actually
sample_data(saliva.prop)$Shannon <- as.vector(as(estimate_richness(saliva.prop, measures="Shannon"),'matrix'))
sample_data(saliva.prop)$Chao1 <- as.vector(as(estimate_richness(saliva.prop, measures="Chao1")[,1],'matrix'))
sample_data(saliva.prop)$InvSimp <- as.vector(as(estimate_richness(saliva.prop, measures="InvSimpson"),'matrix'))
saliva.df <- as(sample_data(saliva.prop), 'data.frame')

## Testing for difference in alpha-diversity between two groups
tapply(saliva.df$Shannon, saliva.df$Group, summary)
with(saliva.df, t.test(Shannon ~ Group)) ## NS

tapply(saliva.df$Chao1, saliva.df$Group, summary)
with(saliva.df, t.test(Chao1~ Group)) ## NS

tapply(saliva.df$InvSimp, saliva.df$Group, summary)
with(saliva.df, t.test(InvSimp ~ Group)) ## NS
## No significant difference in alpha diversity between two groups

## Ordination based on Bray-Curtis and weighted Unifrac 
saliva.wunifrac <- phyloseq::distance(saliva.prop, method="wunifrac")
saliva.bray <- phyloseq::distance(saliva.prop, method="bray")
saliva.ordw <- ordinate(saliva.prop, 'PCoA', saliva.wunifrac)
saliva.ordb <- ordinate(saliva.prop, 'PCoA', saliva.bray)

## Generate a weighted unifrac PCoA plot
plot_scree(saliva.ordw) + xlim(as.character(seq(1,50))) + coord_cartesian(xlim=c(0,50))
plot_ordination(saliva.prop, saliva.ordw, type="samples", color="Group") + 
      geom_point(size=2) + 
      theme_bw() + scale_color_manual(values=c("chocolate1", 'dodgerblue1')) + 
      xlab("PCoA1 (35%)") + ylab("PCoA2 (21%)") + 
      theme(legend.position = "bottom",legend.background = element_rect(color='gray90'))
## no clear separation betweeen cases and control saliva

## Generate a Bray-Curtis PCoA plot
plot_scree(saliva.ordb) + xlim(as.character(seq(1,50))) + coord_cartesian(xlim=c(0,50))
plot_ordination(saliva.prop, saliva.ordb, type="samples", color="Group") + 
      geom_point(size=2) +
      theme_bw() + scale_color_manual(values=c("chocolate1", 'dodgerblue1')) + 
      xlab("PCoA1 (22.4%)") + ylab("PCoA2 (9%)") + 
      theme(legend.position = "bottom",legend.background = element_rect(color='gray90')) 
## no clear separation between case and control saliva
```

Let's check if any covariates could explain the variation in the salivary microbiome
Covariates to include: Group, Age, sex, BMI, oral_dis, smoking_active_2y, alcohol_active_2y. 

```{r}
## Adonis and capscale for wunifrac distance
sal.wuni.perm <- adonis(saliva.wunifrac ~ Group + Age + sex + BMI + oral_dis + smoking_active_2y,
                         data = saliva.df, by='margin', permutations = 9999)
sal.wuni.perm
sal.wuni.cca <- capscale(saliva.wunifrac ~ Group + Age + sex + BMI + oral_dis + smoking_active_2y, 
                         data = saliva.df)
anova(sal.wuni.cca, by='terms', perm=1e4) ## no variables as significant 
anova(sal.wuni.cca, by="margin", perm=1e4) ## no significant variables

## Repeat for bray distance
sal.bray.perm <- adonis2(saliva.bray ~ Group + Age + sex + BMI + oral_dis + smoking_active_2y,
                         data = saliva.df, by='margin', permutations = 9999)
sal.bray.cca <- capscale(saliva.bray ~ Group + Age + sex + BMI + oral_dis + smoking_active_2y, 
                         data = saliva.df)
anova(sal.bray.cca, by='terms', perm=1e4) ## no variables as significant 
anova(sal.bray.cca, by="margin", perm=1e4) ## no signficant variables 
```

The distribution/variance in the salivary microbiome could not be explained by cancer status, age, BMI, oral disease or smoking. These are based on conventional ordination techniques

Next, we examine the compositional data approach, transforming count data into balances, using the isometric log-ratio transformation implemented in the philR package. After such transformation, the transformed count data could be analyzed under the Euclidean space.

```{r}
## Impute zero using the cmultRepl function
otutab.sal.philr <- cmultRepl(t(otu_table(saliva.fil)), method="GBM", output='p-counts')
## output associated tree and metadata from saliva.fil2
tree.sal.philr <- makeNodeLabel(phy_tree(saliva.fil), method='number', prefix='n')
mtdata.sal.philr <- sample_data(saliva.fil)
taxtab.sal.philr <- tax_table(saliva.fil)

gp.sal <- philr(t(otutab.sal.philr), tree.sal.philr, part.weights = 'enorm.x.gm.counts', ilr.weights= 'blw.sqrt')
head(gp.sal)
gp.sal.dist <- dist(gp.sal, method="euclidean")

sal_ord.graph <- plot_ordination(saliva.fil, ordinate(saliva.fil, 'PCoA', distance=gp.sal.dist), type='samples', color="Group") + geom_point(size=1.5) + 
      theme_bw() + scale_color_manual(values=c('springgreen4',"chocolate1")) + 
      xlab("PCoA1 (21.1%)") + ylab("PCoA2 (16.8%)") + stat_ellipse(type="t", linetype=1) + 
      theme(legend.position = "bottom",legend.background = element_rect(color='gray90'))
sal_ord.graph ## Figure 1A
## Again no clear clustering based on group
sal.nomissdata <- which(rownames(gp.sal) %in% rownames(saliva.df))
gp.sal2 <- gp.sal[sal.nomissdata, ]
gp.sal2.dist <- dist(gp.sal2, method="euclidean")
gp.sal.rda <- rda(gp.sal2 ~ Group + Age + sex + BMI + oral_dis + smoking_active_2y, data = saliva.df)
gp.sal.rda
plot(gp.sal.rda)
gp.sal.rda.res <- anova(gp.sal.rda, by="terms", perm=1e5)
gp.sal.rda.res2 <-anova(gp.sal.rda, by="margin", perm=1e5)
gp.sal.rda.res
gp.sal.rda.res2
adonis2(gp.sal2.dist ~ Group + Age + sex + BMI + oral_dis + smoking_active_2y,
                         data = saliva.df, by='margin', permutations = 999)


## This time, testing showed that smoking history in the last two year is borderline significant (p-value < 0.05)
```
Ordination and PERMANOVA showed that only active smoking explains the variance of the salivary microbiome.

Next, we aimed to detect taxa that are differentially abundant between the two group. We will use three approaches: DESeq2 (under negative binomial distribution, designed for sparse RNA-seq data), ANCOMBC (accounting for compositional data structure and sampling bias), and corncob. 

```{r}
# Running ANCOMBC
require(ANCOMBC)
require(microbiome)
## We will use all saliva data, including ones with no associated metadata
saliva.ancom <- saliva.fil
otu_table(saliva.ancom) <- otu_table(abundances(saliva.fil),taxa_are_rows = TRUE)
sample_data(saliva.ancom)$Group <- factor(sample_data(saliva.ancom)$Group, levels=c("control","case"))
saliva.ancom.out <- 
      ancombc(saliva.ancom, formula="Group",p_adj_method = "BH", zero_cut = 0.9, lib_cut=1000,
      group = "Group", struc_zero = TRUE, neg_lb = FALSE, tol=1e-5, max_iter = 300, conserve = FALSE, alpha=0.05)
## we select neg_lb =FALSE since we have <30 samples in the control group, and we expect that the differences between the two group are not large
## Primary results
saliva.ancom.res <- saliva.ancom.out$res
dim(saliva.ancom.out$feature_table) ##  384 taxa among 66 samples were formally tested
## create function to output ancombc results
output_ancombc <- function(physeq, ancom.out, group){
      ancom.res <- ancom.out$res
      if (ncol(ancom.res$diff_abn) == 1) {
            difftx <- rownames(ancom.out$feature_table[which(ancom.res$diff_abn[,group] ==TRUE),])
      } else {
            difftx <- rownames(ancom.res$diff_abn[which(ancom.res$diff_abn[,group] ==TRUE),])
      }
      rs <- data.frame(difftx, ancom.res$W[difftx, group])
      colnames(rs) <- c("taxa", "test_stat")
      rs$qvalue <- ancom.res$q_val[difftx, group]
      rs$coeff <- ancom.res$beta[difftx, group]
      rs$Genus <- tax_table(physeq)[difftx, 'Genus']
      rownames(rs) <- rs$taxa
      rs <- rs[,c(-1)]
      rs <- rs[order(rs$test_stat, decreasing=TRUE),]
      rs$log2fc <- rs$coeff*log2(exp(1))
      return(rs)
}
output_ancombc(physeq = saliva.ancom, ancom.out = saliva.ancom.out, group = "Groupcase")

```

An alternative is to use DESeq2

```{r}
saliva.deseq <- saliva.fil
sample_data(saliva.deseq)$Group <- relevel(as.factor(sample_data(saliva.deseq)$Group), ref='control')
table(rowSums(abundances(saliva.deseq) > 10) > 6)
saliva.deseq <- prune_taxa(rowSums(abundances(saliva.deseq)> 10) >6, saliva.deseq) ## pruning to exclude taxa with very low abundance
summary(sample_sums(saliva.deseq)/sample_sums(saliva.fil)) # median of 95% retention in sample size
length(which(abundances(saliva.deseq) == 0 ))/(ntaxa(saliva.deseq)*nsamples(saliva.deseq)) ## actually moderate number of zero. Sparsity is not that extreme in saliva samples.

## Transform phyloseq into DESeq2 object
ds.sal <- phyloseq_to_deseq2(saliva.deseq, design= ~ Group)
ds.sal <- estimateSizeFactors(ds.sal, type='poscounts')
## Performing testing using DESeq
ds.sal <- DESeq(ds.sal, fitType = "local", test="Wald",sfType="poscounts")
DESeq2::plotMA(ds.sal)
plotDispEsts(ds.sal) ## produced quite a good fit to data
resultsNames(ds.sal)

## create function to summarize DESeq2 output
output_deseq2 <- function(physeq, ds, contrast, alpha, coef, testtype) {
      rs <- results(ds, name=contrast, alpha=alpha, pAdjustMethod ="BH")
      rs_shrink <- lfcShrink(dds=ds, res=rs, coef=coef, type=testtype) ## correcting log2foldchange
      rs_shrink <- rs_shrink[order(rs_shrink$padj, na.last=T),]
      rs_shrink$prank <- seq(1,nrow(rs_shrink))
      rs_shrink$ASV <- rownames(rs_shrink)
      rs_shrink <- as.data.frame(rs_shrink)
      rs_shrink <- rs_shrink[!is.na(rs_shrink$padj),]
      rs_shrink <- rs_shrink[rs_shrink$padj < alpha,]
      rs_shrink$Genus <- tax_table(physeq)[rownames(rs_shrink), "Genus"]
      return(rs_shrink)
}

rs_sal_shrink <- output_deseq2(physeq = saliva.deseq, ds=ds.sal, contrast="Group_case_vs_control", alpha = 0.05, coef=2, testtype = "apeglm")

rs_sal_shrink[rs_sal_shrink$baseMean > 20 & abs(rs_sal_shrink$log2FoldChange) > 0.5,]
## only seq119 (Leptotrichia) is significantly enriched in salivary microbiome of case.
```

Another alternative is to use corncob

```{r}
saliva.deseq
sal.corncob <- differentialTest(formula=~Group, phi.formula = ~Group, formula_null = ~1, phi.formula_null=~1,
                                data=saliva.deseq, taxa_are_rows = FALSE, test="Wald", boot=FALSE,
                                fdr_cutoff = 0.1, fdr="BH") ## set fdr=0.1 for more power
plot(sal.corncob)
sal.corncob$significant_taxa
sal.corncob$p_fdr[sal.corncob$significant_taxa]
sal.corncob$significant_models
## Three taxa are significant. Taxa seq119 (Leptotrichia) and seq433 (Solobacterium) are more abundant in case. 
```

Based on consistency between the three methods, the number of taxa differentiating salivary microbiome of cases and controls is quite minimal. Only seq119 (Leptotrichia) and seq433 (Solobacterium) are more abundant in the CRC saliva.
==========================

Next we construct a correlation network for salivary microbiome alone

```{r}
saliva.fil
saliva.corkeep <- prune_taxa(rowSums(abundances(saliva.fil)>20)>=22, saliva.fil) ## only include taxa with presence in at least 22 samples, and count over 20 in the samples 
saliva.corkeep
summary(sample_sums(saliva.corkeep)/sample_sums(saliva.fil)) ## median of 83% retention

##############################################################################################
#-----------------------------------------------------------------------------------------------------
## Taken from CCLASSO methodology
# The CCLasso function from github
################################################################################
# File: cclasso.R
# Aim : Correlation inference for compositional data through lasso
#-------------------------------------------------------------------------------
# Author : Fang Huaying (Peking University)
# Email  : hyfang@pku.edu.cn
# Date   : 2016-01-08
# Version: 2.0
#-------------------------------------------------------------------------------
# Main function: cclasso(x, counts = FALSE, pseudo = 0.5, k_cv = 3, 
#                        lam_int = c(1e-4, 1), k_max = 20, n_boot = 20) 
#
#  Input:
#           x ------ n x p data matrix (row/column is sample/variable)
#                    n samples & p compositional variables
#      counts ------ Is the compositional data matrix a count matrix? 
#                    Default: FALSE
#      pseudo ------ pseudo count if counts = TRUE
#                    Default: 0.5
#        k_cv ------ folds of cross validation
#                    Default: 3     
#     lam_int ------ tuning parameter interval
#                    Default: [1e-4, 1]
#       k_max ------ maximum iterations for golden section method
#                    Default: 20
#      n_boot ------ Bootstrap times
#                    Default: 20
#  Output: 
#      A list structure contains:
#       var_w ------ variance estimation
#       cor_w ------ correlation estimation
#      p_vals ------ p-values for elements of cor_w equal 0 or not
#      lambda ------ final tuning parameter
#     info_cv ------ information for cross validation
#-------------------------------------------------------------------------------
cclasso <- function(x, counts = FALSE, pseudo = 0.5, k_cv = 3, 
                    lam_int = c(1e-4, 1), k_max = 20, n_boot = 20) {
  n <- nrow(x);
  p <- ncol(x);

  if(counts) {
    x <- x + pseudo;
    x <- x / rowSums(x);
  }
  x <- log(x);
  vx2 <- stats::var(x);

  # Diagonal weight for loss function
  rmean_vx2 <- rowMeans(vx2);
  wd <- 1/diag(vx2 - rmean_vx2 - rep(rmean_vx2, each = p) + mean(rmean_vx2));
  wd2 <- sqrt(wd);
  
  # Some global parameters for optimization with single lambda
  rho <- 1;
  u_f <- eigen(diag(p) - 1/p)$vectors;
  wd_u <- (t(u_f) %*% (wd * u_f))[-p, -p];
  wd_u_eig <- eigen(wd_u);
  d0_wd <- 1 / ( (rep(wd_u_eig$values, each = p-1) + wd_u_eig$values) / 
    (2 * rho) + 1 );
  u0_wd <- wd_u_eig$vectors;

  # Golden section method for the selection of lambda (log10 scale)
  sigma <- vx2;
  lam_int2 <- log10(range(lam_int));
  a1 <- lam_int2[1]; 
  b1 <- lam_int2[2];
  # Store lambda and corresponding cross validation's loss
  lams <- NULL; 
  fvals <- NULL;
  # Two trial points in first 
  a2 <- a1 + 0.382 * (b1 - a1); 
  b2 <- a1 + 0.618 * (b1 - a1);
  fb2 <- cv_loss_cclasso(lambda2 = 10^b2 / rho, x = x, k_cv = k_cv, 
    sigma = sigma, wd = wd, u_f = u_f, u0_wd = u0_wd, d0_wd = d0_wd, 
    wd2 = wd2);
  lams <- c(lams, b2); 
  fvals <- c(fvals, fb2$cv_loss);
  fa2 <- cv_loss_cclasso(lambda2 = 10^a2 / rho, x = x, k_cv = k_cv, 
    sigma = fb2$sigma, wd = wd, u_f = u_f, u0_wd = u0_wd, d0_wd = d0_wd, 
    wd2 = wd2);
  lams <- c(lams, a2); 
  fvals <- c(fvals, fa2$cv_loss);
  # Error tolerance for convergence
  err_lam2 <- 1e-1 * max(1, lam_int2);
  err_fval <- 1e-4;
    
  err <- b1 - a1;
  k <- 0;
  while(err > err_lam2 && k < k_max) {
    fval_max <- max(fa2$cv_loss, fb2$cv_loss);

    if(fa2$cv_loss > fb2$cv_loss) {
      a1 <- a2;      
      a2 <- b2;
      fa2 <- fb2;
      b2 <- a1 + 0.618 * (b1 - a1);
      fb2 <- cv_loss_cclasso(lambda2 = 10^b2 / rho, x = x, k_cv = k_cv, 
        sigma = fa2$sigma, wd = wd, u_f = u_f, u0_wd = u0_wd, d0_wd = d0_wd, 
        wd2 = wd2);

      lams <- c(lams, b2); 
      fvals <- c(fvals, fb2$cv_loss);
    } else {
      b1 <- b2;
      b2 <- a2;
      fb2 <- fa2;
      a2 <- a1 + 0.382 * (b1 - a1);
      fa2 <- cv_loss_cclasso(lambda2 = 10^a2 / rho, x = x, k_cv = k_cv, 
        sigma = fb2$sigma, wd = wd, u_f = u_f, u0_wd = u0_wd, d0_wd = d0_wd, 
        wd2 = wd2);

      lams <- c(lams, a2); 
      fvals <- c(fvals, fa2$cv_loss);
    }
    fval_min <- min(fa2$cv_loss, fb2$cv_loss);      

    k <- k + 1;
    err <- b1 - a1;
    if(abs(fval_max - fval_min) / (1 + fval_min) <= err_fval) {
      break;
    }
  }
  info_cv <- list(lams = lams, fvals = fvals, k = k + 2, 
    lam_int = 10^c(a1, b1)); 
  if(a1 == lam_int2[1] || b1 == lam_int2[2]) {
    cat("WARNING:\n", "\tOptimal lambda is near boundary! ([", 10^a1, ",", 
      10^b1, "])\n", sep = "");
  }

  lambda <- 10^((a2 + b2)/2);
  # Bootstrap for cclasso
  lambda2 <- lambda / rho;
  info_boot <- boot_cclasso(x = x, sigma = fb2$sigma, lambda2 = lambda2, 
    n_boot = n_boot, wd = wd, u_f = u_f, u0_wd = u0_wd, d0_wd = d0_wd);

  return(list(var_w = info_boot$var_w, cor_w = info_boot$cor_w, 
    p_vals = info_boot$p_vals, lambda = lambda, info_cv = info_cv));
}
#-------------------------------------------------------------------------------
# Bootstrap for cclasso
boot_cclasso <- function(x, sigma, lambda2, n_boot = 20, 
                         wd, u_f, u0_wd, d0_wd) {
  n <- nrow(x);
  p <- ncol(x);

  # Store the result of bootstrap
  cors_boot <- matrix(0, nrow = p * (p - 1)/2, ncol = n_boot + 1);
  vars_boot <- matrix(0, nrow = p, ncol = n_boot + 1);
  cors_mat <- matrix(0, p, p);
  ind_low <- lower.tri(cors_mat);

  # Bootstrap procedure
  sam_boot <- matrix(sample(1:n, size = n * n_boot, replace = T), 
    ncol = n_boot);
  for(k in 1:n_boot) {
    ind_samp <- sam_boot[, k];
    sigma2 <- cclasso_sub(sigma = sigma, vx = var(x[ind_samp, ]), 
      lambda2 = lambda2, wd = wd, u_f = u_f, u0_wd = u0_wd, d0_wd = d0_wd);
        
    vars_boot[, k] <- diag(sigma2);
    Is <- 1 / sqrt(vars_boot[, k]);
    cors_mat <- Is * sigma2 * rep(Is, each = p);
    cors_boot[, k] <- cors_mat[ind_low];
  }
  Is <- 1 / sqrt(diag(sigma));
  cors_mat <- sigma * Is * rep(Is, each = p);
  cors_boot[, n_boot + 1] <- cors_mat[ind_low];
  vars_boot[, n_boot + 1] <- diag(sigma);

  #----------------------------------------  
  # Variance estimation via bootstrap
  vars2 <- rowMeans(vars_boot);
  #----------------------------------------
  # Correlations' relationship for artificial null sample
  tol_cor <- 1e-3;
  sam_art0 <- matrix(rnorm(n * p), nrow = n) * rep(sqrt(vars2), each = n);
  cors_art0 <- cor(sam_art0)[ind_low];
  sam_art <- sam_art0 - log(rowSums(exp(sam_art0)));
  sigma_art <- cclasso_sub(sigma = sigma, vx = var(sam_art), 
    lambda2 = lambda2, wd = wd, u_f = u_f, u0_wd = u0_wd, d0_wd = d0_wd);
  Is <- 1 / sqrt(diag(sigma_art));
  cors_mat <- Is * sigma_art * rep(Is, each = p);
  cors_art2 <- cors_mat[ind_low];
  # Bias of estimation between absolute data and cclasso of compositional data
  cors0m2 <- log( ((1 + cors_art0) * (1 - cors_art2)) / ((1 + cors_art2) * 
    (1 - cors_art0)) );
  tmp <- abs(cors_art2) >= tol_cor;
  bias02 <- ifelse(sum(tmp), median(abs(cors0m2)[tmp]), 0);
  # Modification of estimation for cclasso
  cors2 <- log( (1 + cors_boot) / (1 - cors_boot) );    
  cors2mod <- (cors_boot >= tol_cor) * (cors2 + bias02) + 
    (cors_boot <= -tol_cor) * (cors2 - bias02);
  cors2mod <- 1 - rowMeans(2 / (exp(cors2mod) + 1));
  cors2_mat <- diag(p);
  cors2_mat[ind_low] <- cors2mod;
  cors2_mat <- t(cors2_mat);
  cors2_mat[ind_low] <- cors2mod;
  # P-values with null distribution of correlation estimations of absolute data
  p_vals <- pt(cors2mod * sqrt((n - 2) / (1 - cors2mod^2)), df = n - 2);
  p_vals <- ifelse(p_vals <= 0.5, p_vals, 1 - p_vals);
  pval_mat <- diag(p);
  pval_mat[ind_low] <- p_vals;
  pval_mat <- t(pval_mat);
  pval_mat[ind_low] <- p_vals;
  #----------------------------------------

  return(list(var_w = vars2, cor_w = cors2_mat, p_vals = pval_mat));    
}
#-------------------------------------------------------------------------------
# cross validation's loss of cclasso for single lambda
cv_loss_cclasso <- function(lambda2, x, k_cv, sigma, 
                            wd, u_f, u0_wd, d0_wd, wd2) {
  n <- nrow(x);
  p <- ncol(x);

  n_b <- floor(n / k_cv);
  cv_loss <- 0;
  for(k in 1:k_cv) {
    itest <- (n_b * (k - 1) + 1):(n_b * k);
    vxk <- stats::var(x[itest, ]);
    vx2k <- stats::var(x[-itest, ]);

    sigma <- cclasso_sub(sigma = sigma, vx = vx2k, lambda2 = lambda2, 
      wd = wd, u_f = u_f, u0_wd = u0_wd, d0_wd = d0_wd);
    
    dsig <- sigma - vxk;
    tmp <- rowMeans(dsig);
    dsig <- dsig - tmp - rep(tmp, each = p) + mean(tmp);
    cv_loss <- cv_loss + base::norm(wd2 * dsig, "F")^2;
  }

  return(list(cv_loss = cv_loss, sigma = sigma));
}
#-------------------------------------------------------------------------------
# cclasso for single lambda
cclasso_sub <- function(sigma, vx, lambda2, 
                        wd, u_f, u0_wd, d0_wd, 
                        k_max = 200, x_tol = 1e-4) {
  p <- ncol(sigma);
  sigma2 <- sigma;
  LAMBDA <- matrix(0, p, p);
  lambda2 <- matrix(lambda2, p, p);
  diag(lambda2) <- 0;

  k <- 0;
  err <- 1;
  while(err > x_tol && k < k_max) {
    # Update sigma
    x_sigma <- t(u_f) %*% ((sigma2 - vx) - LAMBDA) %*% u_f;
    x_sigma[-p,-p] <- u0_wd %*% ((t(u0_wd) %*% x_sigma[-p, -p] %*% u0_wd) * 
      d0_wd) %*% t(u0_wd);
    sigma_new <- vx + u_f %*% x_sigma %*% t(u_f);
    # Update sigma2
    A <- LAMBDA + sigma_new;
    sigma2_new <- (A > lambda2) * (A - lambda2) + (A < -lambda2) * 
      (A + lambda2);
    # Update Lambda
    LAMBDA <- LAMBDA + (sigma_new - sigma2_new);

    err <- max( abs(sigma_new - sigma)/(abs(sigma) + 1), 
      abs(sigma2_new - sigma2)/(abs(sigma2) + 1) ); 
    k <- k + 1;
    sigma <- sigma_new;
    sigma2 <- sigma2_new;
  }

  if(k >= k_max) {
    cat("WARNING of cclasso_sub:\n", "\tMaximum Iteration:", k_max, 
      "&& Relative error:", err, "!\n");
  }
  
  return(sigma);
}
#-------------------------------------------------------------------------------
#-------------------------------------------------------------------------------
plot_cclasso<- function(cor.map, adj.matrix, phyloseq){
      # Loading library
      require(igraph)
      require(plyr)
      require(RColorBrewer)
      
      # create a graph object from the adjacency matrix
      g <- graph.adjacency(adj.matrix) 
      
      # get the edges as well as the original 
      edgelist <- get.edgelist(g)
      
      edge.cor <- apply(edgelist, MARGIN = 1, FUN = function(x) {cor.map[x[1],x[2]]})
      edgelist <- data.frame(edgelist, edge.cor)
      colnames(edgelist) <- c("from","to", "cor")
      
      # filter edges that link the same node
      filt <- edgelist$from == edgelist$to 
      print("There are some edges that link the same node... now deleting...")
      edgelist <- edgelist[!filt,]
      
      edgelist[,1:2] <- t(apply(edgelist, MARGIN = 1, FUN = function(x){sortEdges(x[1:2])}))
      dupl <- duplicated(paste(edgelist$from, edgelist$to)); sum(dupl)
      edgelist <- edgelist[!dupl,]
      # Take only one direction

      # get the annotation for the nodes that are implicated in the network.
      all.nodes <- unique(c(as.character(edgelist$from), as.character(edgelist$to)))
      # Get the taxonomic classification and abundance of OTUs from original data.
      feat.annot <- taxa_names(phyloseq)[as.numeric(all.nodes)] #OTU names
      feat.annot <- as.data.frame(feat.annot)
      rownames(feat.annot) <- feat.annot$feat.annot
      feat.annot$name <- all.nodes
      feat.annot <- feat.annot[,c(2,1)]
      feat.annot$size <- as.vector(taxa_sums(phyloseq)[as.numeric(all.nodes)]) # abundance
      feat.annot$Order <- as.vector(tax_table(phyloseq)[as.numeric(all.nodes),"Order"]) #Assign taxonomy of the Order level
      print("Finished first part")

      # create the igraph object
      gD <- graph.data.frame(d = edgelist, directed = FALSE, vertices = feat.annot)
      degAll <- igraph::degree(gD, v = V(gD), mode = "all")
      V(gD)$name <- tax_table(phyloseq)[as.numeric(all.nodes), "Genus"]
      #V(gD)$name <- rownames(tax_table(phyloseq)[as.numeric(all.nodes),])
      gD <- set.vertex.attribute(gD, "degree", index = V(gD)$name, value = degAll)
      
      # Calculate betweenness for all nodes
      betAll <- igraph::betweenness(gD, v = V(gD), directed = FALSE) / (((vcount(gD) - 1) * (vcount(gD)-2)) / 2)
      betAll.norm <- (betAll - min(betAll))/(max(betAll) - min(betAll)); rm(betAll)
      
      # Add new node/edge attributes based on the calculated node properties/similarities
      gD <- set.vertex.attribute(gD, "betweenness", index = V(gD)$name, value = betAll.norm)

      ##Calculate edge properties and add to the network
      #Calculate Dice similarities between all pairs of nodes
      dsAll <- similarity.dice(gD, vids = V(gD), mode = "all")
      # The following function will transform a square matrix to an edge driven one and add 
      # values to each edge
      F1 <- function(x) {data.frame(dice = dsAll[which(V(gD)$name == as.character(x$from)), 
                                           which(V(gD)$name == as.character(x$to))])}
      edgelist.ext <- ddply(edgelist, .variables=c("from", "to"), 
                      function(x) data.frame(F1(x))); dim(edgelist.ext)
      gD <- set.edge.attribute(gD, "similarity", index = E(gD), value = 0)
      E(gD)[as.character(edgelist.ext$from) %--% as.character(edgelist.ext$to)]$similarity <- as.numeric(edgelist.ext$dice)
      
      # Edge color based on type of correlations
      E(gD)$color <- ifelse(E(gD)$cor < 0, "blue2", "red") # Red for positive, blue for negative
      # Vertex color based on the Order of the OTU
      V(gD)$color <- c(brewer.pal(8, "Set1"), brewer.pal(11, "Set3"))[as.factor(V(gD)$Order)]

      # resume the network
      print(paste("There are",vcount(gD),"nodes and",ecount(gD),"edges"))
      summary(gD)

      # compute the layout
      l <- layout.auto(gD)
      par(oma=c(2,0.2,0.2,0.2))

      pl <- plot(gD,
            vertex.label = V(gD)$name, 
            vertex.label.cex=0.7, vertex.size=(log10(V(gD)$size))^2/2,
            vertex.label.dist=0.8, vertex.label.color='black',
            edge.arrow.size=2, asp=TRUE, edge.width = abs(E(gD)$cor)*5,
            rescale=TRUE, layout=l
           )
      par(mfrow=c(1,1), fig=c(0,1,0,1), oma=c(2,0,0,0), mar=c(0,0,0,0), new=TRUE)
      ## legend 
      legend(-1,-0.95, legend = levels(factor(tax_table(phyloseq)[as.numeric(all.nodes),"Order"])), 
       fill = c(brewer.pal(8, "Set1"), brewer.pal(11, "Set3")), xpd=TRUE, inset=c(0,0),
      bty='n', cex=0.6,xjust = 0, ncol=5,text.width = 0.4 ,x.intersp = 0.1, title = "Order")
      return(pl)
}
## To beautify the cclasso plot 
sortEdges <- function(var.name = c(node1, node2)){
      if(length(var.name)!=2) stop("Two variable names are needed")
      if(var.name[2]<var.name[1])
      {
            b <- var.name[1]
            var.name[1] <- var.name[2]
            var.name[2] <- b
      }
      return(var.name)
}

##############################################################################################
#-----------------------------------------------------------------------------------------------------

## Apply cclasso to the saliva.corkeep to build correlation network
saliva.cclasso <- cclasso(otu_table(saliva.corkeep), counts=TRUE, k_cv = 3, n_boot = 250)
saliva.cor.map <- saliva.cclasso$cor_w
diag(saliva.cor.map) <- 0 ## all self-interactions are set to 0
saliva.cor.pvalue <- saliva.cclasso$p_vals
saliva.cor.map[which(saliva.cor.pvalue <= 0.01)] ## Lots of interactions are estimated to be significant
### Some exploration
plot(density(as.numeric(saliva.cor.map)), main="All interactions") ## most interaction has strength centering zero
plot(density(as.numeric(saliva.cor.map[abs(saliva.cor.map)>0.55])), main="filtered")

## Building adjacent matrix
saliva.adj.mat <- ((saliva.cor.map >=0.4 | saliva.cor.map <= (-0.5)) & (saliva.cor.pvalue <=0.01)) + 0
## 1 for plotting interactions, 0 for no interaction

## Appply SpiecEasi to the same data
sal.se.mb <- spiec.easi(saliva.corkeep, method='mb', lambda.min.ratio=1e-2, nlambda=20,
                        pulsar.params=list(rep.num=50))
sal.se.gl <- spiec.easi(saliva.corkeep, method="glasso", lambda.min.ratio=1e-2, nlambda=20,
                        pulsar.params=list(rep.num=50))

#sal.spieceasi.mat <- as.matrix(sal.se.mb$refit$stars)
sal.spieceasi.mat <- as.matrix(sal.se.gl$refit$stars)

length(which(sal.spieceasi.mat == 1))
length(which(saliva.adj.mat==1))
## Take the overlapping results between the two methods: cclasso and SpiecEasi 
sal.intersect.cor <- intersect(which(saliva.adj.mat==1), which(sal.spieceasi.mat==1))
length(sal.intersect.cor)
sal.intersect.mat <- matrix(0,nrow=ntaxa(saliva.corkeep), ncol=ntaxa(saliva.corkeep))
sal.intersect.mat[sal.intersect.cor] <- 1

saliva.network <- plot_cclasso(saliva.cor.map, sal.intersect.mat, saliva.corkeep) ## Figure S2
dev.off()
```

Correlation network built for salivary microbiome showed the tight structure of the microbiome. Few significant negative interactions lie with Actinomyces. 

=========================================

We next analyse the gut mucosal microbiome

```{r}
tissue.ps <- prune_samples(sample_data(crc.keep)$sample_type !="saliva", crc.keep)
tissue.ps <- prune_taxa(taxa_sums(tissue.ps) >0, tissue.ps)
tissue.ps
## Explore singleton
tissue.tx.pres <- rowSums(abundances(tissue.ps) > 0)
tissue.singleton <- as.vector(as.matrix(tax_table(tissue.ps)[names(which(tissue.tx.pres == 1)),"Genus"])) ## 480 singleton taxa
names(tissue.singleton) <- rownames(tax_table(tissue.ps)[names(which(tissue.tx.pres == 1)), "Genus"])
sort(table(tissue.singleton),decreasing = TRUE) # Prevotella, Bacteroides, Lachnoclostridium, Blautia, Faecalibacterium, Leptotrichia, etc. 
summary(taxa_sums(tissue.ps)[names(tissue.singleton)]) ## Third quartile is 44 reads.
## Remove singleton taxa with reads < 44
table(tissue.singleton[names(which(taxa_sums(tissue.ps)[names(tissue.singleton)] < 44))]) # covering a wide range of taxa
tissue.tx.rm <- names(which(taxa_sums(tissue.ps)[names(tissue.singleton)] < 44))

tissue.fil <- prune_taxa(!(taxa_names(tissue.ps) %in% tissue.tx.rm), tissue.ps)
summary(sample_sums(tissue.fil)/sample_sums(tissue.ps)) # retained a median 99% of reads 
tissue.fil # 1073 taxa across 136 samples

## use rarefaction curve to assess the library size/coverage 
tissue.ggrare <- ggrare(tissue.fil, step=100, label=NULL, color="sample_type")
ggrare.plot <- tissue.ggrare + xlim(0,10000) + theme(legend.position = 'bottom')
ggrare.plot

### remove samples with low library sizes
sort(sample_sums(tissue.fil)) ## samples P29, P3, P12, P15, P20, B20, P2 All have library size < 1300 reads
tissue.fil <- prune_samples(!(sample_names(tissue.fil) %in% c('P29', 'P3', "P12", "P15", "P20", "B20", "P2")), tissue.fil)
tissue.fil <- prune_taxa(taxa_sums(tissue.fil)>0, tissue.fil)

### Importing a phylogenetic tree to the phyloseq object. Tree was constructed by Iqtree
tissue.tre <- read.newick(file="./tissue_fil_taxa_names_1073seqs_pasta.aln.contree.reroot.newick")
phy_tree(tissue.fil) <- tissue.tre
```

Next we do some exploratory analyses, including ordination 

```{r}
tissue.fil
sample_data(tissue.fil)$oral_dis <- patient.df[match(sample_data(tissue.fil)$Patient, patient.df$Patient), 'oral_dis']
## Remove samples with missing data
tissue.fil2 <- prune_samples(!(is.na(sample_data(tissue.fil)$Age)), tissue.fil)
tissue.fil2 <- prune_taxa(taxa_sums(tissue.fil2) >0, tissue.fil2)
tissue.fil2 ## 120 samples

tissue.median <- median(sample_sums(tissue.fil2)) # 19721
standf.t <- function(x, t=tissue.median) round(t* (x/sum(x)))
tissue.prop <- transform_sample_counts(tissue.fil2, standf.t)

## Basic bar graph to plot abundance of phyla
plot_bar(tissue.prop, fill='Phylum') + geom_bar(aes(color=Phylum, fill=Phylum), stat='identity', position='stack')
plot_richness(tissue.prop, measures=c("Shannon", "Chao1"), color="sample_type")
sample_data(tissue.prop)$Shannon <- estimate_richness(tissue.prop, measures=c("Shannon"))$Shannon
sample_data(tissue.prop)$Chao1 <- estimate_richness(tissue.prop, measures=c("Chao1"))$Chao1
sample_data(tissue.prop)$InvSimpson <- estimate_richness(tissue.prop, measures=c("InvSimpson"))$InvSimpson
tissue.df <- as(sample_data(tissue.prop), "data.frame")

## compare the alpha-diversity between different groups and sample types
tapply(tissue.df$Shannon, tissue.df$sample_type, summary) 
with(tissue.df, kruskal.test(Shannon ~ sample_type)) ## significant p-value
anova(lm(Shannon ~ sample_type, tissue.df)) ## signficant different
TukeyHSD(aov(lm(Shannon ~ sample_type, tissue.df)))
## only non-tumour has significant higher alpha-diversity compared to polyp

tapply(tissue.df$Chao1, tissue.df$sample_type, summary) ## nontumour seems to have the highest diversity
with(tissue.df, kruskal.test(Chao1 ~ sample_type)) # significant p-value
anova(lm(Chao1 ~ sample_type, tissue.df)) # significant
TukeyHSD(aov(lm(Chao1 ~ sample_type, tissue.df)))
## nontumour has significant higher richness compared to polyp and biopsy 
## This may be an artefact of sampling depth 

##### Some basic ordination
tissue.wunifrac <- phyloseq::distance(tissue.prop, method="wunifrac")
tissue.bray <- phyloseq::distance(tissue.prop, method="bray")
tissue.ordw <- ordinate(tissue.prop, 'PCoA', tissue.wunifrac)
tissue.ordb <- ordinate(tissue.prop, 'PCoA', tissue.bray)

plot_scree(tissue.ordw) + xlim(as.character(seq(1,50))) + coord_cartesian(xlim=c(0,50)) + 
      xlab("Principal coordinate") + ylab("Proportion of variance")
plot_ordination(tissue.prop, tissue.ordw, type="samples", color='sample_type') + geom_point(size=1.5) +
      theme_gray() + scale_color_manual(values=c("lightskyblue",'royalblue3',"lightpink1", "indianred4")) + 
      xlab("PCoA1 (34.5%)") + ylab("PCoA2 (17.4%)") + 
      theme(legend.position = "bottom",legend.background = element_rect(color='gray90'))

plot_scree(tissue.ordb) + xlim(as.character(seq(1,50))) + coord_cartesian(xlim=c(0,50)) + 
      xlab("Principal coordinate") + ylab("Proportion of variance")
plot_ordination(tissue.prop, tissue.ordb, type="samples", color='sample_type') + geom_point(size=1.5) +
      theme_gray() + scale_color_manual(values=c("lightskyblue",'royalblue3',"lightpink1", "indianred4")) + 
      xlab("PCoA1 (16.8%)") + ylab("PCoA2 (8.9%)") + 
      theme(legend.position = "bottom",legend.background = element_rect(color='gray90'))

### assess if any covariates explaining the variance in PCoA
tis.wuni.perm <- adonis2(tissue.wunifrac ~ sample_type + Age + sex + BMI + diarrhea_7days + diabetes + HBP + alcohol_active_2y + anatomy, data=tissue.df, by='margin', permutations = 999)
tis.wuni.perm ## sample_type, age, diabetes, HBP 

tis.wuni.cca <- capscale(tissue.wunifrac ~ sample_type + Age + sex + BMI + diarrhea_7days + diabetes + HBP + alcohol_active_2y + anatomy, data=tissue.df)
anova(tis.wuni.cca, by="terms", perm=1e4) # sample_type, age, diabetes, HBP 
anova(tis.wuni.cca, by="margin", perm=1e4) # sample_type, age, diabetes, HBP

## Repeat for Bray distance
tis.bray.perm <- adonis2(tissue.bray ~ sample_type + Age + sex + BMI + diarrhea_7days + diabetes + HBP + alcohol_active_2y + anatomy, data=tissue.df, by='margin', permutations = 999)
tis.bray.perm

tis.bray.cca <- capscale(tissue.bray ~ sample_type + Age + sex + BMI + diarrhea_7days + diabetes + HBP + alcohol_active_2y + anatomy, data=tissue.df)
anova(tis.bray.cca, by="terms", perm=1e4)
anova(tis.bray.cca, by="margin", perm=1e4) 

## Select variables important in constrained analysis
mod0 <- capscale(tissue.wunifrac ~ 1, data=tissue.df)
mod <- step(mod0, scope=formula(tis.wuni.cca), test='perm', direction ='both')
modb <- step(tis.wuni.cca, scope=list(lower=formula(mod0), upper=formula(tis.wuni.cca)), test='perm', direction='both')
# Both forward and reverse model of AIC model selection shows that sample_type, age, diabetes and HBP are significant
```

We next implement the compositional data analysis approach as similar to that applied to salivary microbiome (isometric log-ratio transformation)

```{r}
### Construct ordination for all samples (129, even for samples without covariates)
otutab.tis.philr <- cmultRepl(t(otu_table(tissue.fil)), method='GBM', output='p-counts')
taxtab.tis.philr <- tax_table(tissue.fil)
tree.tis.philr <- makeNodeLabel(as.phylo(phy_tree(tissue.fil)), method='number', prefix='n')

gp.tis <- philr(t(otutab.tis.philr), tree.tis.philr, part.weights = 'enorm.x.gm.counts', ilr.weights='blw.sqrt')
gp.tis[1:10,1:10]
dim(gp.tis)

## Distance based on euclidean space
gp.tis.dist <- dist(gp.tis, method="euclidean")
gp.tis.ord <- ordinate(tissue.fil, formula = ~sample_type, method="PCoA", distance = gp.tis.dist)
plot_scree(gp.tis.ord)

tis_ord.graph <- plot_ordination(tissue.fil, gp.tis.ord, type="samples", color="sample_type") + geom_point(size=1.5)  + 
      theme_bw() + scale_color_manual(values=c("lightskyblue",'royalblue3',"lightpink1", "indianred4")) + 
      xlab("PCoA1 (48.8%)") + ylab("PCoA2 (14.1%)") + stat_ellipse(type="t", linetype=1) + 
      theme(legend.position = "bottom",legend.background = element_rect(color='gray90')) 
tis_ord.graph ## Figure 1B

## Test if any covariates explaining the distribution
## We will test this on a gp.tis2 separately, for only samples with covariates.
otutab.tis2.philr <- cmultRepl(t(otu_table(tissue.fil2)), method='GBM', output='p-counts')
tree.tis2.philr <- makeNodeLabel(as.phylo(phy_tree(tissue.fil2)), method='number', prefix='n')
gp.tis2 <- philr(t(otutab.tis2.philr), tree.tis2.philr, part.weights = 'enorm.x.gm.counts', ilr.weights='blw.sqrt')
gp.tis2.dist <- dist(gp.tis2, method="euclidean")

tis.gp.perm <- adonis2(gp.tis2.dist ~ sample_type + Age + sex + BMI + diarrhea_7days + diabetes + HBP + alcohol_active_2y + anatomy, data=tissue.df, by='margin', permutations = 999)
tis.gp.perm # only sample_type and diabetes are significant.

tis.gp.rda <- rda(gp.tis2 ~ sample_type + Age + sex + BMI + diarrhea_7days + diabetes + HBP + alcohol_active_2y + anatomy, data=tissue.df)
anova(tis.gp.rda, by='terms', perm=1e4) ## sample_type, diabetes
anova(tis.gp.rda, by="margin", perm=1e4) ## sample_type, diabetes
```

These analyses showed that the gut mucosal microbiome between different groups (cases, controls) are significant different. The status of diabetes also likely contributes to the variation in the microbiome observed. 

Examining whether the mucosal microbiome could be clustered

```{r}
## Use the PhILR transformed data for clustering
## prediction strength
tis.gp.pred <- prediction.strength(gp.tis, Gmin = 2, Gmax = 10, M=50, clustermethod = claraCBI, classification="centroid",cutoff=0.5)
tis.gp.pred ## optimal number of cluster is 2 (62%) but low prediction strength

## Gap statistic
gsy = clusGap(t(gp.tis), pam, K.max=8, B=100)
plot_clusgap(gsy) + scale_x_continuous(breaks=c(seq(0,12,2))) ## indicating two groups 
## average silhouette width 
tis.gp.asw <- pamk(gp.tis.dist, krange=2:10, criterion="asw", diss=TRUE, ns=10, critout = TRUE) ## optimal number of clusters is two

ilr.pamclust <- as.factor(pam(x=gp.tis.dist, 2, cluster.only = TRUE, diss=TRUE))
sample_names(tissue.fil) == names(ilr.pamclust) # all TRUE
sample_data(tissue.fil)$CST_ilr <- ilr.pamclust

### Assess the accuracy of membership after clustering by using silhuoette widths
ilr.pamclust.al <- as.array(setNames(as.numeric(sample_data(tissue.fil)$CST_ilr), rownames(sample_data(tissue.fil))))
silwidths <- silhouette((ilr.pamclust.al), dist=gp.tis.dist)
rownames(silwidths) <- names(ilr.pamclust)
plot(silwidths)
print(silwidths)
## For samples with si < (-0.2), consider changing group membership
ilr.pamclust.al["T34"] <- 2
ilr.pamclust.al["T35"] <- 2
ilr.pamclust.al["H34"] <- 2
ilr.pamclust.al["H28"] <- 2
ilr.pamclust.al["T30"] <- 2
ilr.pamclust.al["H30"] <- 2
ilr.pamclust.al["P19"] <- 2
ilr.pamclust.al["P21"] <- 2

#ilr.pamclust.al["T47"] <- 2 
#ilr.pamclust.al["H32"] <- 2
#ilr.pamclust.al["T38"] <- 2
#ilr.pamclust.al["T39"] <- 2

###
plot(silhouette(ilr.pamclust.al, dist=gp.tis.dist))
print(silhouette(ilr.pamclust.al, dist=gp.tis.dist))
## Better now
## Assign membership of ilr.pamclust.al
sample_data(tissue.fil)$CST_ilr <- as.factor(ilr.pamclust.al)

cst.plot <- plot_ordination(tissue.fil, gp.tis.ord, type="samples", color="CST_ilr") + geom_point(size=1.5) + theme_bw() + xlab("PCoA1 (48.8%)") + ylab("PCoA2 (14.1%)") + theme(legend.position = "bottom")
cst.plot ## Figure S1A
## separation of clusters along the first PCoA axis 
#### Examine the membership of the two clusters
t.test(sample_sums(tissue.fil)~as.factor(ilr.pamclust.al))## NS. Group membership is not determined by library size

tissue.df <- as(sample_data(tissue.fil), 'data.frame')
fisher.test(table(tissue.df$Group,ilr.pamclust.al)) ## p-value = 0.002. More controls in group 1. Cases are spread equally between the two groups
table(tissue.df$sample_type, ilr.pamclust.al)

length(which(table(tissue.df$Patient) == 2)) ## 62 pairs of sample (124 samples) from the data
length(which(rowSums(table(tissue.df$Patient, ilr.pamclust.al) == 2 )==1))
## 56/62 pairs of patient samples (90.3%) belong to the same CST cluster. 

### Randomforest to identify taxa most differentiating betweewn the two CSTs
rdf.df <- as.data.frame(gp.tis)
identical(rownames(rdf.df), names(ilr.pamclust.al)) # TRUE
rdf.df$CST_ilr <- as.factor(ilr.pamclust.al)
rdf.rf <- rfsrc(CST_ilr ~., data=rdf.df, ntree=10000, block.size = 50, importance = 'permute')
print(rdf.rf)
## Overall OOB is ~ 10.8%. 
plot.rfsrc(rdf.rf)
## export VIMP from randomforest object
class.vimp <- vimp.rfsrc(rdf.rf)
## VIMP for all 
sort(class.vimp$importance[,1], decreasing = TRUE)[1:20] ## n20, n15, n17
## VIMP important for cluster 1
sort(class.vimp$importance[,2], decreasing = TRUE)[1:20] ## n20, n15, n17, n3, n75
## VIMP important for cluster 2
sort(class.vimp$importance[,3], decreasing = TRUE)[1:20] ## n20, n15, n73, n3, 
## Extract information from tree balance
info_balance_tissue <- function(node) {
      bal <- name.balance(tree.tis.philr, taxtab.tis.philr[,-c(7)], node, return.votes=c('up', 'down'))
      return(bal)
}
info_balance_tissue('n17') # Proteobacteria/Actinobacteria 
info_balance_tissue("n20") # GammaProteobacteria (E. coli/Sutterella)/Campilobacterota+Desulfobacterota
info_balance_tissue("n15") # Actinobacteria + Proteobacteria/Firmicutes + Fusobacteria 
info_balance_tissue("n3") # other phylums/Lachnospiraceae
info_balance_tissue("n73") # E. coli/Morganella

## Visualizing the value of gp.tis in these balance
gp.tis.subset <- convert_to_long(gp.tis, get_variable(tissue.fil, 'CST_ilr')) %>% filter(coord %in% c('n15', "n17", 'n20', 'n73', 'n3'))
balance.plot <- ggplot(gp.tis.subset, aes(x=labels, y=value)) + geom_boxplot(fill='lightgrey') + 
      facet_grid(.~coord, scale='free_x') + 
      xlab('CST_ILR') + ylab('Balance value') + theme_bw()
### Use this as supplementary figure
require(gridExtra)
grid.arrange(cst.plot, balance.plot, nrow=1, widths =c(0.5, 0.5)) ## Figure S1

## Investigate which covariates could contribute to the differentiation between CST 1 and 2. 
patient.df2 <- as(sample_data(tumbiop.keep),'data.frame')
patient.df2 <- patient.df2[!(is.na(patient.df2$Age)),]
dim(patient.df2)
table(patient.df2$Group)

rownames(patient.df2) <- patient.df2$Patient 
#patient.df2 <- patient.df2[patient.df2$Group=='case',]
## some investigation
with(patient.df2, t.test(Age ~ CST_ilr)) ## NS
with(patient.df2, t.test(BMI ~ CST_ilr)) ## NS
fisher.test(with(patient.df2, table(Group, CST_ilr))) ## borderline significant
fisher.test(with(patient.df2, table(diabetes, CST_ilr))) ## NS
fisher.test(with(patient.df2, table(diarrhea_7days, CST_ilr))) # NS
chisq.test(with(patient.df2, table(HBP, CST_ilr))) # NS 
chisq.test(with(patient.df2, table(sex, CST_ilr))) ## NS
chisq.test(with(patient.df2, table(oral_dis, CST_ilr))) # NS
chisq.test(with(patient.df2, table(alcohol_active_2y, CST_ilr))) # NS
chisq.test(with(patient.df2, table(alcohol_history, CST_ilr))) #NS 
```

The gut mucosal microbiome could be split into two clusters, matching separation along the first PCoA. Cluster 1 is more enriched in GammaProteobacteria (E. coli), as cluster 2 is more enriched in Actinobacteria (Collinsella) and Lachnospiraceae. The ratio of Proteobacteria/Actinobacteria is the most different. 

We next plot a heatmap to visualize the relative abundance of most prevalent taxa. We will agglomerate ASVs of the same genus into one for easy summary and visualization
```{r}
## First we correct some taxonomic assignment
tax_table(tissue.fil)[which(tax_table(tissue.fil)[,"Family"] == "Parvimonas"), "Genus"] <- "Parvimonas"

tissue.txglom <- tax_glom(tissue.fil, taxrank = "Genus",NArm = FALSE)
tissue.txglom.dt <- microbiome::transform(tissue.txglom, transform='compositional')
tax_table(tissue.txglom.dt)['seq96', 'Genus'] <- 'unc_Lachnospiraceae'
tissue.core <- core_members(tissue.txglom.dt, detection =1/100, prevalence=15/100)
length(tissue.core)
tissue.txglom.dt.fil <- prune_taxa(tissue.core, tissue.txglom.dt)
summary(sample_sums(tissue.txglom.dt.fil)/sample_sums(tissue.txglom.dt)) 
## 24 genera summarize a median of 85% data in tissue microbiome
tax_table(tissue.txglom.dt.fil)[tissue.core,]

## Prepare for heatmap plot 
## Reorder the sample
cst1_order <- ilr.pamclust.al[which(ilr.pamclust.al == 1)]
cst1_order <- cst1_order[c(which(str_match(names(cst1_order),"B") == "B"), which(str_match(names(cst1_order),"P") == "P"),which(str_match(names(cst1_order),"H") == "H"), which(str_match(names(cst1_order),"T") == "T"))]
cst2_order <- ilr.pamclust.al[which(ilr.pamclust.al == 2)]
cst2_order <- cst2_order[c(which(str_match(names(cst2_order),"B") == "B"), which(str_match(names(cst2_order),"P") == "P"),which(str_match(names(cst2_order),"H") == "H"), which(str_match(names(cst2_order),"T") == "T"))]

s.order <- names(c(cst1_order, cst2_order))
s.order
tax_table(tissue.txglom.dt.fil)['seq1','Genus'] <- 'Escherichia'
tax_table(tissue.txglom.dt.fil)['seq29', "Genus"] <- 'Ruminococcus torques'
tax_table(tissue.txglom.dt.fil)['seq54', "Genus"] <- "Ruminococcus gnavus"

M.taxa.order <- c("Escherichia", "Klebsiella", "Bacteroides", "Parabacteroides", "Prevotella", "Alloprevotella", 
                  "Collinsella", "Sutterella", "Megamonas", "Phascolarctobacterium", "Ruminococcus torques", "Ruminococcus gnavus",
                  "Bifidobacterium", "Faecalibacterium", "Megasphaera", "Blautia", "unc_Lachnospiraceae", "Dorea", 
                  "Fusobacterium", "Streptococcus", "Leptotrichia","Gemella", "Parvimonas", "Peptostreptococcus")
new.taxa.order <- rev(taxa_names(tissue.txglom.dt.fil)[match(M.taxa.order, tax_table(tissue.txglom.dt.fil)[,'Genus'])])

### Plotting
library("wesanderson")
pal <- wes_palette("Zissou1", 5, type='discrete')
taxa.col <- rev(c(rep('red3',2), rep("blue3",4), 'gold4', 'red3', rep('forestgreen',4), 'gold4', rep('forestgreen',5),
                  'darkviolet', 'forestgreen', 'darkviolet', rep('forestgreen',3)))

plot_heatmap.2 <- function(ps, sample.label=NULL, taxa.label=NULL, ...) {
      hm <- plot_heatmap(ps, taxa.label="Genus", sample.order= s.order, taxa.order = new.taxa.order)
      hm <- hm + theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
                       axis.text.y=element_text(size=10, face='italic', colour = taxa.col))
      #low = "#190a0a"; high = "#ff6666"; trans = scales::log_trans(4); na.value = "black" # From plot_heatmap defaults
      new_gradient <- scale_fill_gradientn(colours=c('gray10',pal),
                                           values=c(0,0.01, 0.15, 0.3, 0.6, 1), na.value = NA, guide="colourbar", name="Relative\nproportion")
      hm <- hm + theme(plot.margin=unit(c(0,0.5,0.5,0.5),"lines"))
      hm <- hm + new_gradient
      hm <- hm + geom_tile(colour='black') #
      hm <- hm + ylab("Taxa")
      hm$layers <- hm$layers[2] #
      return(hm)
}

hm <- plot_heatmap.2(tissue.txglom.dt.fil, taxa.label = "Genus") +
      theme(legend.position = 'right')
hm
```

Next we annotate the sample type and CST information on the heatmap

```{r}
make_hcb <- function(data, var, name = NULL, fillScale = NULL, ...) {
      hcb <- ggplot(data=data, aes_string(x="index", y=1, fill=var)) + 
            geom_raster() +
            scale_y_continuous(expand=c(0,0), breaks=1, labels=name) + 
            scale_x_continuous(expand=c(0,0)) +
            xlab(NULL) + ylab(NULL) +
            theme(axis.title=element_blank(), axis.ticks=element_blank()) +
            theme(axis.text.x=element_blank()) +
            theme(axis.text.y=element_text(size=12, face="bold")) +
            theme(plot.margin=unit(c(0,0,0,0),"lines")) +
            theme(axis.text=element_text(margin=unit(0,'null'))) +
            guides(fill=F)
      if(!is.null(fillScale)) hcb <- hcb + fillScale
      return(hcb)
}
tissue.df2 <- as(sample_data(tissue.txglom.dt.fil), 'data.frame')
tissue.df2 <- tissue.df2[s.order, ]
tissue.df2$index <- seq(1, nsamples(tissue.txglom.dt.fil))

cst_hcb <- make_hcb(tissue.df2, 'CST_ilr', name='CST', fillScale = scale_fill_manual(values=c("gray80", 'gray40')))
cst_hcb <- cst_hcb + annotate('text', x=tapply(tissue.df2$index, tissue.df2[,'CST_ilr', drop=T], mean), y =1, label=levels(tissue.df2$CST), size=6)
cst_hcb

sample_hcb <- make_hcb(tissue.df2, 'sample_type', name='Sample', fillScale = scale_fill_manual(values=c('biopsy'='lightskyblue', 'nontumour'='lightpink1','polyp' = 'royalblue3', 'tumour'='indianred4')))
sample_hcb

mush <- function(hmap, hcbs) {
      require(gtable)
      require(grid)
      require(gridExtra)
      cbgs <- lapply(hcbs, ggplotGrob)
      hmg <- ggplotGrob(hmap)
      
      # Make sure both plots have the same dimensions in grob objects
      for (i in seq_along(cbgs)) {
            cbgs[[i]] <- gtable_add_cols(cbgs[[i]], widths=unit(1,"null"), pos=8)
            cbgs[[i]] <- gtable_add_cols(cbgs[[i]], widths=unit(1,"null"), pos=8)
      }
      ## Make sure that both plots have the same widths
      cbWidths <- lapply(cbgs, function(x) x$widths)
      maxWidth <- do.call(unit.pmax, cbWidths)
      maxWidth <- unit.pmax(hmg$widths, maxWidth)
      ## replace widths in each grob object with maxWidth
      hmg$widths <- maxWidth
      for (i in seq_along(cbgs)){
            cbgs[[i]]$widths <- maxWidth
      }
      heights <- unit.c(unit(rep(1,length(cbgs)), "lines"), unit(1, "null"))
      rval <- do.call(arrangeGrob, args = c(cbgs, list(hmg), ncol=1, heights=list(heights)))
      return(rval)
}

big.hm <- mush(hm, list(sample_hcb, cst_hcb))
vis <- grid.arrange(big.hm) ## Figure 1D
```

We next investigate the beta-diversity among the samples, using the transformed philr values

```{r}
## Pairwise beta-diversity within and between patients
tis.ilr.mat <- as.matrix(gp.tis.dist)
pID <- tissue.df[,c("Group", "Patient")]
pID$ID <- rownames(pID)
pID <- pID[match(rownames(tis.ilr.mat), pID$ID),]
pNO <- levels(as.factor(pID$Patient))
patientid.notpaired <- names(which(table(pID$Patient) == 1)) ## ID that are not paired
pNO <- pNO[!(pNO %in% patientid.notpaired)]

for (i in 1:length(pNO)) {print(length(pID[which(pID$Patient == pNO[i]),1]))} ## all have 2 samples in one pair.
Y <- matrix(rep(NA,length(pNO)),nrow=length(pNO), ncol=1)
for (i in 1:length(pNO)){
      coord <- combn(pID[which(pID$Patient==pNO[i]),3],2)
      Y[i,1] <- tis.ilr.mat[coord[1], coord[2]]
      print(paste('Successfully for patient ID ', i, collapse=""))
}
Y <- as.data.frame(Y)
rownames(Y) <- pNO
colnames(Y) <- "distance"

Y$Group <- pID[match(rownames(Y), pID$Patient),'Group']
ggplot(Y, aes(Group, distance)) + geom_boxplot()
summary(Y[Y$Group=='case','distance']) ## median of pairwise distance in the CRC patients (tumour and healthy tissue) is 4.72
summary(Y[Y$Group=='control','distance']) ## median of pairwise distance in the control (biopsy and polyp) is 5.7
wilcox.test(distance ~ Group, data=Y) ## not significant difference

Y$tis1 <- rownames(tissue.df[match(rownames(Y), tissue.df$Patient),])
Y$tis2 <- Y$tis1
Y$tis2 <- str_replace(str_replace(Y$tis2, "H", "T"), "B", "P")
tissue_sample_sums <- sample_sums(tissue.fil)
Y$tis1_readsum <- tissue_sample_sums[match(Y$tis1, names(tissue_sample_sums))]
Y$tis2_readsum <- tissue_sample_sums[match(Y$tis2, names(tissue_sample_sums))]
Y$min_readsum <- apply(Y[,c("tis1_readsum", "tis2_readsum")],1, min)
cor(Y$distance, Y$min_readsum, method = "spearman") ## some moderate association 
ggplot(Y, aes(min_readsum, distance)) + geom_point() + geom_smooth()
## There is a weak trend of linear relationship between the sequencing coverage for each sample and the difference in beta-diversity

## Pairwise difference for samples belonging to the same sample_type
tumour.mat.id <- which(colnames(tis.ilr.mat) %in% Y[Y$Group == "case", 'tis2'])
tumour.distmat <- tis.ilr.mat[tumour.mat.id, tumour.mat.id]
tumour.pwdist <- tumour.distmat[which(lower.tri(tumour.distmat, diag=FALSE))]
summary(tumour.pwdist) ## median pairwise difference between non-paired tumour samples is 12.5

nontm.mat.id <- which(colnames(tis.ilr.mat) %in% Y[Y$Group == 'case', 'tis1'])
nontm.distmat <- tis.ilr.mat[nontm.mat.id, nontm.mat.id]
nontm.pwdist <- nontm.distmat[which(lower.tri(nontm.distmat, diag=FALSE))]
summary(nontm.pwdist) ## median pairwise difference between non-paired healthy samples is 12.08

biop.mat.id <- which(str_detect(colnames(tis.ilr.mat),'B'))
biop.distmat <- tis.ilr.mat[biop.mat.id,biop.mat.id]
biop.pwdist <- biop.distmat[which(lower.tri(biop.distmat, diag=FALSE))]
summary(biop.pwdist) ## median pairwise difference between non-paired biopsy samples is 11.506

polyp.mat.id <- which(str_detect(colnames(tis.ilr.mat), "P"))
polyp.distmat <- tis.ilr.mat[polyp.mat.id, polyp.mat.id]
polyp.pwdist <- polyp.distmat[which(lower.tri(polyp.distmat, diag=FALSE))]
summary(polyp.pwdist) ## median pairwise difference between non-paired polyp samples is 11.574

## pairwise difference for salivary microbiome
sal.ilr.mat <- as.matrix(gp.sal.dist)
sal_conid <- which(str_match(rownames(sal.ilr.mat), "S0") == "S0")
salcon.distmat <- sal.ilr.mat[sal_conid, sal_conid]
salcon.pwdist <- salcon.distmat[which(lower.tri(salcon.distmat, diag = FALSE))]
summary(salcon.pwdist) ## median is 32 (quite high)

sal_caseid <- which(str_match(rownames(sal.ilr.mat), "S1") == "S1")
salcase.distmat <- sal.ilr.mat[sal_caseid, sal_caseid]
salcase.pwdist <- salcase.distmat[which(lower.tri(salcase.distmat, diag = FALSE))]
summary(salcase.pwdist) ## median is 36.53 (quite high)

wilcox.test(salcon.pwdist, salcase.pwdist) ## significant p-value < 0.05
t.test(sample_sums(saliva.fil) ~ sample_data(saliva.fil)$Group) ## NS, while no difference in library size.

########### Drawing into one plot ##############
wcon.dist <- Y[Y$Group =='control','distance']
wcase.dist <- Y[Y$Group=='case', 'distance']
H <- list(wcon.dist, wcase.dist, biop.pwdist, polyp.pwdist, nontm.pwdist, tumour.pwdist)
Hm <- melt(H)
colnames(Hm) <- c('value', 'comparison')
table(Hm$comparison)
Hm$comparison <- str_replace(Hm$comparison, "1", "within_control")
Hm$comparison <- str_replace(Hm$comparison, "2", "within_case")
Hm$comparison <- str_replace(Hm$comparison, "3", "between_biopsy")
Hm$comparison <- str_replace(Hm$comparison, "4", "between_polyp")
Hm$comparison <- str_replace(Hm$comparison, "5", "between_nontumour")
Hm$comparison <- str_replace(Hm$comparison, "6", "between_tumour")
Hm$comparison <- factor(Hm$comparison, levels = c('within_control', 'within_case', 'between_biopsy', 'between_polyp','between_nontumour', 'between_tumour'))

pwdist.graph <- ggplot(Hm, aes(comparison, value, fill=comparison)) + geom_boxplot() + 
      theme_bw() + 
      scale_fill_manual(values=c("gray80", "gray60","lightskyblue",'royalblue3',"lightpink1","indianred4")) + 
      theme(axis.text.x=element_text(angle=45, size=9, vjust=1, hjust=1 ), legend.position = 'none') + 
      xlab("") + ylab("Beta-diversity")
pwdist.graph ## Figure 1C
## Anova and TukeyHSD to see signficant difference among the comparisons
TukeyHSD(aov(lm(value ~ comparison, Hm)))
```

Now we combine ordination figures, pairwise distance and heatmap 

```{r}
library(egg)
library(gridExtra)
yy <- egg::ggarrange(sal_ord.graph, tis_ord.graph, pwdist.graph, widths = c(0.33, 0.33, 0.34))
yy
fig1 <- grid.arrange(yy, big.hm, nrow=2, heights =c(0.4, 0.6)) ## Figure 1
ggsave(fig1, filename="./ordination_relabund_heatmap.pdf", height=8.5, width=11, units='in', dpi=600)
```
=====================================================
Differential abundance analyses
Identify taxa significant abundant between these comparisons
1/ Paired tumour and nontumour (CRC patients)
2/ Paired biopsy and polyp (control)
3/ Tumour and biopsy (case vs control)
4/ Cancer stage III-IV and stage II 

1/ Paired tumour and nontumour
```{r}
### DESeq2 to compare paired tumour and nontumour tissue 
tissuecase.deseq <- prune_samples(sample_data(tissue.fil)$Group == 'case', tissue.fil)
tissuecase.deseq <- prune_taxa(taxa_sums(tissuecase.deseq) > 0, tissuecase.deseq)
table(rowSums(abundances(tissuecase.deseq) > 5) >= 4)
tissuecase.keep <- prune_taxa(rowSums(abundances(tissuecase.deseq) > 5) > 4, tissuecase.deseq)
summary(sample_sums(tissuecase.keep)/sample_sums(tissuecase.deseq)) ## median of 93% retention

length(which(abundances(tissuecase.keep) ==0))/(nsamples(tissuecase.keep)*ntaxa(tissuecase.keep)) ## 80% of cells are zero. This data is extremely sparse even after filtering. 
## Consider applying the zero-inflated NB model 
sample_data(tissuecase.keep)$sample_type ## nontumour is reference level
ds_crc_zinbwave <- phyloseq_to_deseq2(tissuecase.keep, design = ~0+Patient + sample_type) ## design without intercept
ds_crc_zinbwave <- zinbwave(ds_crc_zinbwave, epsilon=1e10, verbose=TRUE, K=0, observationalWeights=TRUE,
                            BPPARAM=BiocParallel::SerialParam(), X="~1") 
assay(ds_crc_zinbwave, "weights")[1:5, 1:5]
dds_crc_zinb <- DESeqDataSet(ds_crc_zinbwave, design=~0+Patient + sample_type)
dds_crc_zinb <- estimateSizeFactors(dds_crc_zinb, type="poscounts")
## Use size factor normalization by poscounts

dds_crc_zinb <- DESeq(dds_crc_zinb, test="LRT", reduced=~Patient, sfType="poscounts", minmu=0.5,
                      minReplicatesForReplace = 7, fitType = "local", betaPrior = FALSE, useT=TRUE)
## Observe dispersion plot
DESeq2::plotMA(dds_crc_zinb)
plotDispEsts(dds_crc_zinb) ## better fit compared to poscount sizefactors without zero-inflation 
## When the counts are very small (for some taxa), it becomes very difficult to accurately estimate the dispersion
## Refitting the dispersion by removing low abundance taxa
## This will filter taxa solely for the dispersion trend estimation, but still use a larger dataset for the rest of the analysis
keepForDispTrend <- rowSums(DESeq2::counts(dds_crc_zinb) >=10)>=8
dds2_crc <- estimateDispersionsFit(dds_crc_zinb[keepForDispTrend,])
plotDispEsts(dds2_crc) ## remove dispersion fit for very low counts

dispersionFunction(dds_crc_zinb) <- dispersionFunction(dds2_crc)
dds_crc_zinb <- estimateDispersionsMAP(dds_crc_zinb)
dds_crc_zinb <- nbinomLRT(dds_crc_zinb, reduced=~Patient, minmu=0.5)

## Extract results from DESeq2 analysis
resultsNames(dds_crc_zinb)

rs_crc_shrink <- output_deseq2(physeq = tissuecase.deseq,ds = dds_crc_zinb, contrast="sample_typetumour", alpha=0.05, coef=44, testtype="ashr")

up.tumour <- rs_crc_shrink[rs_crc_shrink$log2FoldChange > 0 & rs_crc_shrink$baseMean > 20, c(1,2,5,8)]
up.tumour
down.tumour <- rs_crc_shrink[rs_crc_shrink$log2FoldChange < 0 & rs_crc_shrink$baseMean > 20, c(1,2,5,8)]
down.tumour

##########################
## Apply a different approach 
### ANCOMBC: compositional data analysis approach with correction for sampling bias
tissuecase.ancom <- tissuecase.deseq
otu_table(tissuecase.ancom) <- otu_table(abundances(tissuecase.ancom),taxa_are_rows = TRUE)

case.ancom.out <- ancombc(phyloseq = tissuecase.ancom, formula="Patient + sample_type", p_adj_method = "holm", zero_cut=0.90, lib_cut=1000, group = "sample_type", struc_zero=TRUE, neg_lb=TRUE, tol=1e-5, max_iter=300, conserve=TRUE, alpha=0.05, global=FALSE)

## Primary results
case.ancom.res <- case.ancom.out$res
length(which(case.ancom.res$q_val[,'sample_typetumour'] < 0.05)) ## 52 taxa have adjusted pvalue < 0.05
length(which(case.ancom.res$diff_abn[,'sample_typetumour'] ==TRUE)) ## 52 taxa are differential abundant

case.ancom.rs <- output_ancombc(tissuecase.ancom, ancom.out = case.ancom.out, group="sample_typetumour")

ancom.up.tumour <- case.ancom.rs[case.ancom.rs$test_stat >= 1 & case.ancom.rs$coeff> 0,]

ancom.down.tumour <-case.ancom.rs[case.ancom.rs$test_stat < (-1) & case.ancom.rs$coeff < (0),]

## Consistent results output from both DESeq2 and ANCOMBC
up.tumour.cons <- intersect(rownames(ancom.up.tumour), rownames(up.tumour))
up.tumour[up.tumour.cons,] ## 11 taxa 

down.tumour.cons <- intersect(rownames(ancom.down.tumour), rownames(down.tumour))
down.tumour[down.tumour.cons,] ## only 5 taxa

## Plotting for visualization
## Use ANCOMBC results table for consistent plotting
diff.tumour <- rbind(ancom.up.tumour[up.tumour.cons,], ancom.down.tumour[down.tumour.cons,])
diff.tumour['seq227', 'Species'] <- "Hungatella"
diff.tumour['seq67', "Species"] <- "Campylobacter rectus"
diff.tumour['seq40', "Species"] <- "Selenomonas sputigena"
diff.tumour['seq46', "Species"] <- "Fusobacterium nucleatum"
diff.tumour['seq13', "Species"] <- "Leptotrichia "
diff.tumour['seq3', "Species"] <- "Collinsella aerofaciens"
diff.tumour['seq58', "Species"] <- "Blautia"
diff.tumour['seq14', "Species"] <- "Bacteroides vulgatus"
diff.tumour["seq261", "Species"] <- "Bacteroides nordii"
diff.tumour["seq240", "Species"] <- "Osillibacter ruminantium"
diff.tumour["seq88", "Species"] <- "Streptococcus sanguinis" # oral (100% to HOMD)
diff.tumour["seq15", "Species"] <- "Gemella" # oral
diff.tumour["seq210", "Species"] <- "Lachnoclostridium"
diff.tumour["seq61", "Species"] <- "Fusobacterium"
diff.tumour["seq157", "Species"] <- "Dorea formicigenerans"
diff.tumour["seq126", "Species"] <- "Parabacteroides merdae"

## which one is oral
diff.tumour$type <- c(0,0,1,0,1,1,1,1,1,0,1,0,0,0,0,0)
## Order the seq based on log2foldChange
diff.tumour <- diff.tumour[order(diff.tumour$log2fc, decreasing=TRUE),]
species.level <- diff.tumour[order(diff.tumour$log2fc, decreasing=TRUE), "Species"]
diff.tumour$Species <- factor(diff.tumour$Species, levels=unique(species.level))

## Set margin
par(mar=c(0,0,0,0))
par(oma=c(0,0,0,0))
diffcrc.graph <- ggplot(diff.tumour, aes(Species, log2fc, fill=factor(type))) + geom_bar(stat="identity") + coord_flip() +
      ylab("(paired tumour/non-tumour)") + xlab("Species") + theme_bw() + 
      theme(axis.text.x=element_text(angle=0, size=12, vjust=1, hjust=1 ), legend.position = 'none') + 
      scale_fill_manual(values=c("gray9", "lightcoral"), labels=c("no", "yes"), guide=guide_legend(title="Oral microbiome", label.position="bottom"))
diffcrc.graph ## Figure 2A
```

2/ Paired biopsy and polyp

```{r}
con.deseq <- prune_samples(sample_data(tissue.fil)$Group == 'control', tissue.fil)
con.deseq <- prune_taxa(taxa_sums(con.deseq) > 0, con.deseq)
table(str_extract(sample_names(con.deseq), "[0-9]+"))
# Patient 12, 15, 2, 29 and 3 will be removed since they are not paired
con.deseq <- prune_samples(!(sample_names(con.deseq) %in% c("B12", "B15", "B2", "B3", "B29")), con.deseq)
con.deseq <- prune_taxa(taxa_sums(con.deseq) >0, con.deseq)
con.deseq # 19 pairs of controls for comparison

table(rowSums(abundances(con.deseq) > 5) > 2)
con.keep <- prune_taxa(rowSums(abundances(con.deseq) > 5) > 2, con.deseq)
summary(sample_sums(con.keep)/sample_sums(con.deseq)) ## median of 96% retention

length(which(abundances(con.keep) ==0))/(nsamples(con.keep)*ntaxa(con.keep)) ## 78% of cells are zero. Data is extremely sparse
sample_data(con.keep)$sample_type ## biopsy is reference level

## Consider applying the zero-inflated NB model 

ds_con_zinbwave <- phyloseq_to_deseq2(con.keep, design = ~0+Patient + sample_type) ## design without intercept
ds_con_zinbwave <- zinbwave(ds_con_zinbwave, epsilon=1e10, verbose=TRUE, K=0, observationalWeights=TRUE,
                            BPPARAM=BiocParallel::SerialParam(), X="~1") 
assay(ds_con_zinbwave, "weights")[1:5, 1:5]
dds_con_zinb <- DESeqDataSet(ds_con_zinbwave, design=~0+Patient + sample_type)
dds_con_zinb <- estimateSizeFactors(dds_con_zinb, type="poscounts")

dds_con_zinb <- DESeq(dds_con_zinb, test="LRT", reduced=~Patient, sfType="poscounts", minmu=0.5,
                      minReplicatesForReplace = 7, fitType = "local", betaPrior = FALSE, useT=TRUE)
## Observe dispersion plot
DESeq2::plotMA(dds_con_zinb)
plotDispEsts(dds_con_zinb) ## dispersion trend match well

resultsNames(dds_con_zinb)
rs_con_shrink <- output_deseq2(physeq=con.deseq, ds = dds_con_zinb, contrast = "sample_typepolyp",alpha = 0.05,coef = 20, testtype = "ashr")
rs_con_shrink ## seq76 (Faecalibacterium), seq95 and seq30 (Prevotella) are less abundant in polyp

## Apply ANCOMBC to paired control samples
con.ancom <- con.deseq
otu_table(con.ancom) <- otu_table(abundances(con.ancom),taxa_are_rows = TRUE)
con.ancom.out <- ancombc(phyloseq = con.ancom, formula="Patient + sample_type", p_adj_method = "holm", zero_cut=0.90, lib_cut=1000, group = "sample_type", struc_zero=TRUE, neg_lb=FALSE, tol=1e-5, max_iter=300, conserve=FALSE, alpha=0.05)

## Primary results from paired control comparison
con.ancom.res <- con.ancom.out$res
length(which(con.ancom.res$q_val[,'sample_typepolyp'] < 0.05)) ## 5 taxa have adjusted pvalue < 0.05
length(which(con.ancom.res$diff_abn[,'sample_typepolyp'] ==TRUE))

con.ancom.rs <- output_ancombc(physeq=con.ancom, ancom.out=con.ancom.out, group="sample_typepolyp")

con.ancom.rs

## Consistent results between DESeq2 and ANCOMBC only showed seq76 (Faecalibacterium) is significantly reduced in polyps. 
rs_con_shrink["seq76",]
```

3/ Compare tumour and biopsies microbiome

```{r}
tumbiop.ps <- prune_samples(sample_data(tissue.fil)$sample_type == "tumour" | 
                                  sample_data(tissue.fil)$sample_type =="biopsy", tissue.fil)
tumbiop.ps <- prune_taxa(taxa_sums(tumbiop.ps) > 0, tumbiop.ps)

tumbiop.keep <- prune_taxa(rowSums(abundances(tumbiop.ps) > 5) > 4, tumbiop.ps)
tumbiop.keep
summary(sample_sums(tumbiop.keep)/sample_sums(tumbiop.ps)) ## median of 90% retention

length(which(abundances(tumbiop.keep) ==0))/(nsamples(tumbiop.keep)*ntaxa(tumbiop.keep)) ## 77% of cells are zero. Data is extremely sparse
sample_data(tumbiop.keep)$sample_type ## biopsy is the reference level
## 
ds_tis <-  phyloseq_to_deseq2(tumbiop.keep, design =~sample_type)
ds_tis <- estimateSizeFactors(ds_tis, type="poscounts") 
## Don't apply the zero-inflated model in case that the two groups has clear differences (presence/absence) dependent on actual zero counts

ds_tis <- DESeq(ds_tis, test="Wald", minmu=0.5,
                      minReplicatesForReplace = 7, fitType = "local", betaPrior = FALSE, useT=TRUE)

## Observe dispersion plot
DESeq2::plotMA(ds_tis)
plotDispEsts(ds_tis) 

resultsNames(ds_tis)
rs_tis_shrink <- output_deseq2(physeq=tumbiop.keep,ds = ds_tis,contrast="sample_type_tumour_vs_biopsy", alpha=0.05,coef = 2, testtype="apeglm")

uptum.dsrs <- rs_tis_shrink[rs_tis_shrink$log2FoldChange > 0 & rs_tis_shrink$baseMean > 20, c(1,2,5,8)]
uptum.dsrs ## F. nucleatum, Selenomonas, Leptotrichia, etc. 
rs_tis_shrink[rs_tis_shrink$log2FoldChange < 0 & rs_tis_shrink$baseMean > 20, c(1,2,5,8)] ## seq135, Fusobacterium mortiferum

### Let's try ancombc
tumbiop.ancom <- tumbiop.ps
otu_table(tumbiop.ancom) <- otu_table(abundances(tumbiop.ancom),taxa_are_rows = TRUE)
tumbiop.ancom.out <- ancombc(phyloseq = tumbiop.ancom, formula="sample_type", p_adj_method = "holm", zero_cut=0.90, lib_cut=1000, group = "sample_type", struc_zero=TRUE, neg_lb=TRUE, tol=1e-5, max_iter=300, conserve=TRUE, alpha=0.05)

## Primary results from paired control comparison
tumbiop.ancom.res <- tumbiop.ancom.out$res
length(which(tumbiop.ancom.res$q_val[,'sample_typetumour'] < 0.05)) ## 93 taxa have adjusted pvalue < 0.05
length(which(tumbiop.ancom.res$diff_abn[,'sample_typetumour'] ==TRUE))

tumbiop.ancom.rs <- output_ancombc(physeq = tumbiop.ancom,ancom.out = tumbiop.ancom.out,group = "sample_typetumour")
uptum.ancom <- tumbiop.ancom.rs[which(tumbiop.ancom.rs$test_stat > 1 & tumbiop.ancom.rs$coeff > 0),] ## Gemella, Peptostreptococcus, Fusobacterium
upbiops.ancom <- tumbiop.ancom.rs[which(tumbiop.ancom.rs$test_stat < (-1) & tumbiop.ancom.rs$coeff < 0),] ## F. mortiferum (seq8), seq96 (unc_Lachno)

uptum.ancom
upbiops.ancom

## Try out corncob 
tumbiop.keep
tumbiop.corncob <- differentialTest(formula=~sample_type, phi.formula=~1, formula_null = ~1, phi.formula_null = ~1,
                                    data=tumbiop.keep, test="LRT", boot=FALSE, fdr_cutoff=0.05, fdr="BH", taxa_are_rows=FALSE)
plot(tumbiop.corncob)
tumbiop.corncob$significant_taxa
tumbiop.corncob$discriminant_taxa_DA

## Consistent results between DESeq2, ancombc, corncob: 
# overabundant in tumour microbiomes:
intersect(rownames(uptum.dsrs), rownames(uptum.ancom)) # seq15, seq49, seq46, seq13, seq67, seq40
intersect(rownames(uptum.dsrs), tumbiop.corncob$significant_taxa) # seq15, seq49
intersect(rownames(upbiops.ancom), tumbiop.corncob$significant_taxa) # seq96, seq258, seq8, seq698

## Use log2foldchange output from ancombc
diff.tumbiop <- rbind(uptum.ancom, upbiops.ancom)
diff.tumbiop <- diff.tumbiop[c("seq15", "seq49", "seq46", "seq13", "seq67", "seq40", "seq96", "seq258", "seq8"), ]
diff.tumbiop["seq15", "Species"] <- "Gemella"
diff.tumbiop["seq49", "Species"] <- "Peptostreptococcus stomatis"
diff.tumbiop["seq46", "Species"] <- "Fusobacterium nucleatum"
diff.tumbiop["seq13", "Species"] <- "Leptotrichia"
diff.tumbiop['seq67', "Species"] <- "Campylobacter rectus"
diff.tumbiop['seq40', "Species"] <- "Selenomonas sputigena"
diff.tumbiop["seq96", "Species"] <- "unc_Lachnospiraceae"
diff.tumbiop["seq258", "Species"] <- "Sutterella"
diff.tumbiop["seq8", "Species"] <- "Fusobacterium mortiferum"

## which one is oral
diff.tumbiop$type <- c(1,1,1,1,1,1,0,0,0)
## Order the seq based on log2foldChange
diff.tumbiop <- diff.tumbiop[order(diff.tumbiop$log2fc, decreasing=TRUE),]
species.level <- diff.tumbiop[order(diff.tumbiop$log2fc, decreasing=TRUE), "Species"]
diff.tumbiop$Species <- factor(diff.tumbiop$Species, levels=unique(species.level))

## Set margin
par(mar=c(0,0,0,0))
par(oma=c(0,0,0,0))
difftumbiop.graph <- ggplot(diff.tumbiop, aes(Species, log2fc, fill=factor(type))) + geom_bar(stat="identity") + coord_flip() +
      ylab("(tumour/biopsy)\nlog2FoldChange") + theme_bw() + 
      theme(axis.text.x=element_text(angle=0, size=12, vjust=1, hjust=1 ), legend.position = 'bottom') + 
      scale_fill_manual(values=c("gray9", "lightcoral"), labels=c("no", "yes"), guide=guide_legend(title="Oral microbiome", label.position="bottom"))
difftumbiop.graph # Figure 2A

```

4/ Between cancer stages (only in tumour samples first)

```{r}
## Try ANCOMBC first
tissuecase.deseq
sample_data(tissuecase.deseq)$C_stage_short <- NA
stageII <- which(sample_data(tissuecase.deseq)$Cancer_stage %in% c("IIA", "IIB"))
stageIII <- which(sample_data(tissuecase.deseq)$Cancer_stage %in% c("IIIA", "IIIB", "IIIC", "IV"))
sample_data(tissuecase.deseq)[stageII, "C_stage_short"] <- "II"
sample_data(tissuecase.deseq)[stageIII, "C_stage_short"] <- "IIIandIV"
sample_data(tissuecase.deseq)$C_stage_short <- factor(sample_data(tissuecase.deseq)$C_stage_short)
Cstage.deseq <- prune_samples(!(is.na(sample_data(tissuecase.deseq)$C_stage_short)), tissuecase.deseq)
Cstage.deseq <- prune_samples(sample_data(Cstage.deseq)$sample_type == "tumour", Cstage.deseq)
Cstage.deseq <- prune_taxa(taxa_sums(Cstage.deseq)> 0, Cstage.deseq)
Cstage.deseq

Cstage.ancom <- Cstage.deseq
otu_table(Cstage.ancom) <- otu_table(abundances(Cstage.ancom),taxa_are_rows = TRUE)
stage.ancom.out <- ancombc(phyloseq=Cstage.ancom, formula="C_stage_short", p_adj_method = "BH", zero_cut = 0.9, lib_cut=1000, group="C_stage_short", struc_zero=TRUE, neg_lb=TRUE, tol=1e-5, max_iter=300, conserve=FALSE, alpha=0.05)

stage.ancom.res <- stage.ancom.out$res
length(which(stage.ancom.res$q_val[,'C_stage_shortIIIandIV'] < 0.05)) ## 70 taxa have adjusted pvalue < 0.05
length(which(stage.ancom.res$diff_abn[,'C_stage_shortIIIandIV'] ==TRUE)) ## 70 taxa are potentially differential abundant

stage.ancom.rs <- output_ancombc(physeq=Cstage.ancom,ancom.out = stage.ancom.out,group="C_stage_shortIIIandIV")
upstage3.ancom <- stage.ancom.rs[stage.ancom.rs$test_stat > 0 & stage.ancom.rs$log2fc > 0,]
upstage2.ancom <- stage.ancom.rs[stage.ancom.rs$test_stat < 0 & stage.ancom.rs$log2fc < 0 ,]

## Try DESeq2 
Cstage.deseq
Cstage.keep <- prune_taxa(rowSums(abundances(Cstage.deseq)> 0) > 2, Cstage.deseq)
summary(sample_sums(Cstage.keep)/sample_sums(Cstage.deseq)) ## mean 90% of sample_sums retention
Cstage.keep
sample_data(Cstage.keep)$C_stage_short
fisher.test(table(sample_data(Cstage.keep)$CST_ilr, sample_data(Cstage.keep)$C_stage_short)) ## NS
length(which(abundances(Cstage.keep) == 0))/(ntaxa(Cstage.keep)*nsamples(Cstage.keep)) ## 80% of cells are zero, sparse.

### Apply the zero-inflated model 
ds.crcstage.zinb <- phyloseq_to_deseq2(Cstage.keep, design = ~C_stage_short)
ds.crcstage.zinb <- zinbwave(ds.crcstage.zinb, epsilon=1e10, verbose=TRUE, K=0, observationalWeights=TRUE, BPPARAM=BiocParallel::SerialParam(), X="~1")
dds_stage_zinb <- DESeqDataSet(ds.crcstage.zinb, design = ~C_stage_short)
ds.crcstage <- dds_stage_zinb
ds.crcstage <- estimateSizeFactors(ds.crcstage, type="poscount")

ds.crcstage <- DESeq(ds.crcstage, test="LRT", reduced=~1, minmu=0.5,
                      minReplicatesForReplace = 7, fitType = "local", betaPrior = FALSE, useT=TRUE)
taxa_names(Cstage.keep)[which(mcols(ds.crcstage)$weightsFail)] ## 

## Observe dispersion plot
DESeq2::plotMA(ds.crcstage)
plotDispEsts(ds.crcstage) ## dispersion estimate plot fits well

resultsNames(ds.crcstage)
rs_crcstage_shrink <- output_deseq2(physeq=Cstage.keep, ds=ds.crcstage, contrast="C_stage_short_IIIandIV_vs_II", alpha=0.05, coef=2, testtype="ashr")
upstage3.rsds <- rs_crcstage_shrink[rs_crcstage_shrink$log2FoldChange > 0 & rs_crcstage_shrink$baseMean > 20, c(1,2,5,8)] ## seq53 and seq13
downstage3.rsds <- rs_crcstage_shrink[rs_crcstage_shrink$log2FoldChange < 0 & rs_crcstage_shrink$baseMean > 20, c(1,2,5,8)] 

### Try out for corncob 
Cstage.keep
crcstage.corncob <- differentialTest(formula=~C_stage_short, phi.formula=~C_stage_short, formula_null=~1, phi.formula_null=~C_stage_short, 
                                     data=Cstage.keep, test="Wald", boot=FALSE, fdr_cutoff = 0.1, fdr="BH", taxa_are_rows = FALSE)
plot(crcstage.corncob)
### No significant results

### Look for consistent results
intersect(rownames(upstage3.rsds), rownames(upstage3.ancom)) # seq13 (Leptotrichia) 
intersect(rownames(downstage3.rsds), rownames(upstage2.ancom)) # seq166 (Prevotella)
```

5/ Repeat the comparison between cancer stages on on nontumour tissue

```{r}
tissuecase.deseq
Nstage.deseq <- prune_samples(!(is.na(sample_data(tissuecase.deseq)$C_stage_short)), tissuecase.deseq)
Nstage.deseq <- prune_samples(sample_data(Nstage.deseq)$sample_type == "nontumour", Nstage.deseq)
Nstage.deseq <- prune_taxa(taxa_sums(Nstage.deseq)> 0, Nstage.deseq)
Nstage.deseq

Nstage.ancom <- Nstage.deseq
otu_table(Nstage.ancom) <- otu_table(abundances(Nstage.ancom),taxa_are_rows = TRUE)
Nstage.ancom.out <- ancombc(phyloseq=Nstage.ancom, formula="C_stage_short", p_adj_method = "holm", zero_cut = 0.9, lib_cut=1000, group="C_stage_short", struc_zero=TRUE, neg_lb=TRUE, tol=1e-5, max_iter=300, conserve=TRUE, alpha=0.05)

Nstage.ancom.res <- Nstage.ancom.out$res
length(which(Nstage.ancom.res$q_val[,'C_stage_shortIIIandIV'] < 0.05)) ## 77 taxa have adjusted pvalue < 0.05
length(which(Nstage.ancom.res$diff_abn[,'C_stage_shortIIIandIV'] ==TRUE))

Nstage.ancom.rs <- output_ancombc(physeq=Nstage.ancom, ancom.out = Nstage.ancom.out, group="C_stage_shortIIIandIV")
upNstage3.ancom <- Nstage.ancom.rs[Nstage.ancom.rs$test_stat > 0 & Nstage.ancom.rs$log2fc > 0,]
downNstage3.ancom <- Nstage.ancom.rs[Nstage.ancom.rs$test_stat < 0 & Nstage.ancom.rs$log2fc < 0 ,]

## Try DESeq2 
Nstage.deseq
Nstage.keep <- prune_taxa(rowSums(abundances(Nstage.deseq)> 0) > 2, Nstage.deseq)
summary(sample_sums(Nstage.keep)/sample_sums(Nstage.deseq)) ## mean 90% of sample_sums retention
Nstage.keep
sample_data(Nstage.keep)$C_stage_short
fisher.test(table(sample_data(Nstage.keep)$CST_ilr, sample_data(Nstage.keep)$C_stage_short)) ## NS
length(which(abundances(Nstage.keep) == 0))/(ntaxa(Nstage.keep)*nsamples(Nstage.keep)) ## 79% of cells are zero, sparse.

### Apply the zero-inflated model 
ds.Nstage.zinb <- phyloseq_to_deseq2(Nstage.keep, design = ~C_stage_short)
ds.Nstage.zinb <- zinbwave(ds.Nstage.zinb, epsilon=1e10, verbose=TRUE, K=0, observationalWeights=TRUE, BPPARAM=BiocParallel::SerialParam(), X="~1")
ds.Nstage <- DESeqDataSet(ds.Nstage.zinb, design = ~C_stage_short)
ds.Nstage <- estimateSizeFactors(ds.Nstage, type="poscount")

ds.Nstage <- DESeq(ds.Nstage, test="LRT", reduced=~1, minmu=0.5,
                      minReplicatesForReplace = 7, fitType = "local", betaPrior = FALSE, useT=TRUE)
taxa_names(Nstage.keep)[which(mcols(ds.Nstage)$weightsFail)] ## 

## Observe dispersion plot
DESeq2::plotMA(ds.Nstage)
plotDispEsts(ds.Nstage) ## dispersion estimate plot fits well

resultsNames(ds.Nstage)
rs_Nstage_shrink <- output_deseq2(physeq=Nstage.keep, ds=ds.Nstage, contrast="C_stage_short_IIIandIV_vs_II", alpha=0.05, coef=2, testtype="ashr")

upNstage3.dsrs <- rs_Nstage_shrink[rs_Nstage_shrink$log2FoldChange > 0 & rs_Nstage_shrink$baseMean > 20, c(1,2,5,8)] ## seq53 (Fusobacterium)
downNstage3.dsrs <- rs_Nstage_shrink[rs_Nstage_shrink$log2FoldChange < 0 & rs_Nstage_shrink$baseMean > 20, c(1,2,5,8)] # no taxa

intersect(rownames(upNstage3.ancom), rownames(upNstage3.dsrs)) ## no taxa
```

The comparison showed that only seq13 (Leptotrichia) is deemed as more abundant in tumour of advanced cancer stage III-IV, as compared to stageII. Comparing in cancer stages between nontumour samples yielded no significant overlapping taxa.

Next we visualize the proportional abundance of selected group of bacteria between cancer stages

```{r}
tissue.dt <- transform_sample_counts(tissue.fil2, function(x) x/sum(x))
sample_sums(tissue.dt)

Fuso.tx <- rownames(tax_table(tissue.dt)[which(tax_table(tissue.dt)[,'Genus'] == "Fusobacterium"),])
## Assign species identification to Fusobacterium if blast identity > 99%
Fuso.df <- as.data.frame(sample_data(tissue.dt))
identical(rownames(Fuso.df),sample_names(tissue.dt)) # TRUE
Fuso.df$F_nucleatum <- sample_sums(prune_taxa(c("seq46", "seq53", "seq68", "seq79", "seq80", "seq130", "seq153", "seq186", "seq347", "seq373",
             "seq480", "seq723", "seq1059", "seq1284"), tissue.dt))
Fuso.df$F_mortiferum <- sample_sums(prune_taxa(c("seq8", "seq27", "seq135", "seq168", "seq294", "seq296", "seq337", "seq389",
                                                 "seq424", "seq607", "seq634", "seq645", "seq697", 'seq1159'), tissue.dt))
Fuso.df$Fusobacterium <- sample_sums(prune_taxa(Fuso.tx, tissue.dt))

## Leptotrichia
Lepto.tx <- rownames(tax_table(tissue.dt)[which(tax_table(tissue.dt)[,'Genus'] == "Leptotrichia"),])
Fuso.df$Leptotrichia <- sample_sums(prune_taxa(Lepto.tx, tissue.dt))

## Collinsella as the opposite trend
Collinsella.tx <- rownames(tax_table(tissue.dt)[which(tax_table(tissue.dt)[,'Genus'] == "Collinsella"),])
Fuso.df$Collinsella <- sample_sums(prune_taxa(Collinsella.tx, tissue.dt))

##### SummarySE function
summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE) {
      require(plyr)
      
      # New version of length which can handle NA's: if na.rm==T, don't count them
      length2 <- function (x, na.rm=FALSE) {
            if (na.rm) sum(!is.na(x))
            else       length(x)
      }
      
      # This does the summary. For each group's data frame, return a vector with
      # N, mean, and sd
      datac <- ddply(data, groupvars, .drop=.drop,
                     .fun = function(xx, col) {
                           c(N    = length2(xx[[col]], na.rm=na.rm),
                             mean = mean  (xx[[col]], na.rm=na.rm),
                             sd   = sd     (xx[[col]], na.rm=na.rm)
                           )
                     },
                     measurevar
      )
      
      # Rename the "mean" column    
      datac <- plyr::rename(datac, c("mean" = measurevar))
      
      datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean
      
      # Confidence interval multiplier for standard error
      # Calculate t-statistic for confidence interval: 
      # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
      ciMult <- qt(conf.interval/2 + .5, datac$N-1)
      datac$ci <- datac$se * ciMult
      return(datac)
}

### Visualization in different cancer stages
table(Fuso.df$Cancer_stage)/2
crc.df <- Fuso.df[which(Fuso.df$Group == 'case'),]
crc.df$Cancer_stage_short <- NA
crc.df[which(crc.df$Cancer_stage %in% c("IIA", "IIB")), "Cancer_stage_short"] <- "II"
crc.df[which(crc.df$Cancer_stage %in% c("IIIA", "IIIB", "IIIC", "IV")), "Cancer_stage_short"] <- "III-IV"
crc.df$Cancer_stage_short

B <- melt(crc.df, id.vars=c('sample_type', 'Cancer_stage_short'),
          measure.vars=c("F_mortiferum", "F_nucleatum", "Leptotrichia", "Collinsella"))
colnames(B) <- c("sample_type", "Cancer_stage_short", "taxa", "value")
Bc <- summarySE(B, measurevar = 'value', groupvars = c("sample_type", "Cancer_stage_short", 'taxa'))
relab.graph <- ggplot(Bc, aes(x=taxa, y=value, fill=Cancer_stage_short)) + geom_bar(position=position_dodge(), stat='identity', width=0.7) + 
      geom_errorbar(aes(ymin=value-se, ymax=value+se), width=0.3, position=position_dodge(0.7)) +  
      scale_fill_manual(name="Cancer stage", values=c("lightgoldenrod2", "goldenrod4")) + xlab("Taxa") + 
      facet_wrap(~sample_type, ncol=1) + 
      ylab("Relative abundance") + theme_bw() + 
      theme(legend.position = "bottom", axis.text = element_text(size=11,colour='black'), axis.title.x = element_blank(),
            axis.title.y=element_text(size=12, colour='black', face='bold'), panel.grid.minor = element_blank(),
            panel.grid.major.x = element_blank(), axis.text.x = element_text(angle=30, hjust=1)) 
relab.graph ## Figure 2C
```

==============
Combine graph 

```{r}
gg <- egg::ggarrange(diffcrc.graph, difftumbiop.graph, heights = c(0.65, 0.35))

fig2 <- grid.arrange(gg, relab.graph, nrow=1, widths =c(0.65, 0.35)) ## figure 2 
ggsave(fig2, filename="./Diff_abundance.pdf", height=8.5, width=11, units='in', dpi=600)
```

====================
Construct a correlation network for gut mucosal microbiome in cancer patients

```{r}
tissuecase.deseq
table(rowSums(abundances(tissuecase.deseq) > 10) >=15)
case.fil <- prune_taxa(rowSums(abundances(tissuecase.deseq) > 10) >=15, tissuecase.deseq)
summary(sample_sums(case.fil)/sample_sums(tissuecase.deseq)) ## median 76% of total reads are retained.
tax_table(case.fil)[,"Genus"]
mm <- cmultRepl(t(otu_table(case.fil)), method="GBM", output='p-counts')
otu_table(case.fil) <- otu_table(t(mm), taxa_are_rows = FALSE)

case.cclasso <- cclasso(otu_table(case.fil), counts=TRUE, k_cv=3, n_boot=250,pseudo = 0)
case.cor.map <- case.cclasso$cor_w
diag(case.cor.map) <- 0 ## all self-interation is 0 
case.cor.pvalue <- case.cclasso$p_vals
case.cor.map[which(case.cor.pvalue <= 0.01)] # A lot of interactions are significant

## Visualization
plot(density(as.numeric(case.cor.map)), main="All relations")
plot(density(as.numeric(case.cor.map[abs(case.cor.map) > 0.4])), main='filtered')

length(which(case.cor.pvalue <=0.01 & abs(case.cor.map) >=0.4))## 210 interactions
## building adjacent matrix
## For easy visualization, we retained the interactions with highest correlation scores
case.adj.mat <- ((case.cor.map >=0.37 | case.cor.map <= (-0.395)) & case.cor.pvalue <=0.01) + 0

#### Run the same dataset with SpiecEasi
case.se.mb <- spiec.easi(case.fil, method='mb', lambda.min.ratio=1e-2, nlambda=20,
                         pulsar.params=list(rep.num=50))
case.se.gl <- spiec.easi(case.fil, method="glasso", lambda.min.ratio=1e-2, nlambda=20,
                         pulsar.params=list(rep.num=50))

case.se.mb
case.spieceasi.mat <- as.matrix(case.se.mb$refit$stars)

## Investigate the correlation score for significant SpiecEasi correlations
case.se.corstrength <- case.cor.map[which(case.spieceasi.mat == 1 )]
summary(unique(case.se.corstrength[case.se.corstrength > 0])) #176/2 interactions ## median 0.4495
summary(unique(case.se.corstrength[case.se.corstrength < 0]))

#case.adj.mat <- as.matrix(case.se.mb$refit$stars)
## 
length(which(case.spieceasi.mat == 1)) ## 232 interactions
length(which(case.adj.mat == 1)) ## 262 interactions
which(case.spieceasi.mat == 1)
which(case.adj.mat == 1)

case.intersect.cor <- intersect(which(case.spieceasi.mat==1),which(case.adj.mat==1))
length(case.intersect.cor) # 178 interactions are consistent between the two methods

#setdiff(which(case.spieceasi.mat==1), case.intersect.cor)

case.intersect.mat <- matrix(0,nrow=ntaxa(case.fil), ncol=ntaxa(case.fil))
case.intersect.mat[case.intersect.cor] <- 1

#casetissue.network <- plot_cclasso(case.cor.map, case.adj.mat, case.fil)
casetissue.network <- plot_cclasso(case.cor.map, case.intersect.mat, case.fil)
#casetissue.network <- plot_cclasso(case.cor.map, case.spieceasi.mat, case.fil)
dev.off()

#####
## Oral clusters in gut also present in saliva?
gut_oral1 <- c("seq13", "seq40", "seq46", "seq60", "seq25", "seq15", "seq49", "seq24")
salcor.temp <- saliva.cor.map
dim(salcor.temp)
colnames(salcor.temp) <- taxa_names(saliva.corkeep)
rownames(salcor.temp) <- taxa_names(saliva.corkeep)
gut_oral1.id <- match(gut_oral1, colnames(salcor.temp))
gut_oral1.id <- gut_oral1.id[!(is.na(gut_oral1.id))]
salcor.temp[gut_oral1.id, gut_oral1.id]
saliva.cor.pvalue[gut_oral1.id, gut_oral1.id]
## significant hits include seq40-seq46, seq46-seq25, seq25-seq15, seq49-seq15

## oral cluster 2
gut_oral2 <- c("seq2", "seq9", "seq12", "seq17", "seq75")
gut_oral2.id <- match(gut_oral2, colnames(salcor.temp))
salcor.temp[gut_oral2.id, gut_oral2.id]
saliva.cor.pvalue[gut_oral2.id, gut_oral2.id]
## significant hits include seq2-seq75, seq12-seq17, 
```

Construct a correlation network but only for control patients

```{r}
con.deseq
table(rowSums(abundances(con.deseq) > 10) >=6)
con.fil <- prune_taxa(rowSums(abundances(con.deseq) > 10) >=6, con.deseq)
summary(sample_sums(con.fil)/sample_sums(con.deseq)) ## median 88% of total reads are retained.
tax_table(con.fil)[,"Genus"]

nn <- cmultRepl(t(otu_table(con.fil)), method="GBM", output='p-counts')
otu_table(con.fil) <- otu_table(t(nn), taxa_are_rows = FALSE)
con.fil

con.cclasso <- cclasso(otu_table(con.fil), counts=TRUE, k_cv = 3, n_boot = 250, pseudo=0)
con.cor.map <- con.cclasso$cor_w
diag(con.cor.map) <- 0 ## all self-interation is 0 
con.cor.pvalue <- con.cclasso$p_vals
con.cor.map[which(con.cor.pvalue <= 0.01)] # A lot of interactions are significant

## Visualization
plot(density(as.numeric(con.cor.map)), main="All relations")
plot(density(as.numeric(con.cor.map[abs(con.cor.map) > 0.5])), main='filtered')

length(which(con.cor.pvalue <=0.01 & abs(con.cor.map) >=0.5))## 196 interactions
## building adjacent matrix
con.adj.mat <- ((con.cor.map >=0.4 | con.cor.map <= (-0.45)) & con.cor.pvalue <=0.01) + 0

## Run SpiecEasi on the same dataset
con.se.mb <- spiec.easi(con.fil, method='mb', lambda.min.ratio=1e-2, nlambda=20,
                         pulsar.params=list(rep.num=50))
con.se.gl <- spiec.easi(case.fil, method="glasso", lambda.min.ratio=1e-2, nlambda=20,
                         pulsar.params=list(rep.num=50))
con.spieceasi.mat <- as.matrix(con.se.mb$refit$stars)

## 
length(which(con.spieceasi.mat == 1))
length(which(con.adj.mat==1))
con.intersect.cor <- intersect(which(con.adj.mat==1), which(con.spieceasi.mat==1))
length(con.intersect.cor)
con.intersect.mat <- matrix(0,nrow=ntaxa(con.fil), ncol=ntaxa(con.fil))
con.intersect.mat[con.intersect.cor] <- 1

contissue.network <- plot_cclasso(con.cor.map, con.intersect.mat, con.fil)
```

Tumour-associated ASVs. How prevalent are they in the gut and saliva.

```{r}
taxaset1 <- rownames(diff.tumour[which(diff.tumour$log2fc > 0 & diff.tumour$type==1),])
tumour.ps <- prune_samples(sample_data(tissue.fil)$sample_type == "tumour", tissue.fil)
tumour.ps <- prune_taxa(taxa_sums(tumour.ps)> 0, tumour.ps)

colMeans(apply(abundances(tumour.ps)[taxaset1,],1 , function(x) x/sample_sums(tumour.ps)))
## mean abundances range from 0.007 (Campylobacter) to 0.046 (Leptotrichia seq13)
prevalence(t(apply(abundances(tumour.ps)[taxaset1,],1 , function(x) x/sample_sums(tumour.ps))), count=FALSE)
## Prevalence is quite ok: 25% of tumours had seq13 Leptotrichia, while 34% tumours had seq46 (F. nucleatum)

## Account for all other F. nucleatum and Leptotrichia ASVs
length(which(Fuso.df[Fuso.df$sample_type == 'tumour', c("F_nucleatum")] > 0.001 ))/42 ## 62% present, 54% has relative abundance > 0.1%
length(which(Fuso.df[Fuso.df$sample_type == 'tumour', c("Leptotrichia")] > 0.001))/42 ## 52% present, 40% has relative abundance > 0.1%

taxaset2 <- rownames(diff.tumour[which(diff.tumour$log2fc > 0 & diff.tumour$type==0),])
colMeans(apply(abundances(tumour.ps)[taxaset2,],1 , function(x) x/sample_sums(tumour.ps)))
## much lower abundance, but range from 0.0015 to 0.0023)
prevalence(t(apply(abundances(tumour.ps)[taxaset2,],1 , function(x) x/sample_sums(tumour.ps))), count=FALSE)
## Hungatella (seq227) present in 34% of tumours, Lachnoclostridium (seq210) present in 41.8% of tumours

## looking for samples with low abundance of tumour-associated ASVs (<1%)
length(which(rowSums(apply(abundances(tumour.ps)[c(taxaset1, taxaset2),], 1, function(x) x/sample_sums(tumour.ps))) >= 0.01))

crc_sus.id <- names(which(rowSums(apply(abundances(tumour.ps)[c(taxaset1),], 1, function(x) x/sample_sums(tumour.ps))) < 0.01))
length(crc_sus.id)
nontumour.sus.id <- str_replace(crc_sus.id, "T", "H")
Fuso.df[crc_sus.id,]
Fuso.df[nontumour.sus.id,]
for (x in 1:length(crc_sus.id)) {
      print(crc_sus.id[x])
      print(tax_table(tumour.ps)[names(sort(abundances(tumour.ps)[,crc_sus.id[x]],decreasing=TRUE)[1:8]),"Genus"])
      print((sort(abundances(tumour.ps)[,crc_sus.id[x]],decreasing=TRUE)[1:8])/sample_sums(tumour.ps)[crc_sus.id[x]])
}

for (x in 1:length(nontumour.sus.id)) {
      print(nontumour.sus.id[x])
      print(tax_table(tissue.fil)[names(sort(abundances(tissue.fil)[,nontumour.sus.id[x]],decreasing=TRUE)[1:8]),"Genus"])
      print((sort(abundances(tissue.fil)[,nontumour.sus.id[x]],decreasing=TRUE)[1:8])/sample_sums(tissue.fil)[nontumour.sus.id[x]])
}

tax_table(tumour.ps)[names(sort(abundances(tumour.ps)[,"T18"],decreasing=TRUE)[1:8]),"Genus"]
### The odd eggs. Tumour samples with dominance of other taxa 
## T11: Klebsiella/Enterobacteriaceae (>90%); H11: Klebsiella + Enterobacteriacaea (90%)
# T13: E.coli (87%);  H13: E. coli as well (96%)
# T17: Streptococcus + E. coli (80%); H17: E. coli + Collinsella 
# T20: Megamonas;  H20: Megamonas (27%)
# T21: Enterococcus (17%); H21: Bacteroides + Enterococcus 
# T25: F. mortiferum and F. varium (30%); H25: Collinsella + F. mortiferum 
# T26: Megamonas (36%); H26: Megamonas (16%)
# T31: E. coli (37%) + Megamonas (9%); H31: Collinsella + Megamonas
# T33: Megamonas; H33: Collinsella + Megamonas
# T34: F. mortiferum (12%) + E. coli (11%); H34: E. coli (15%)
# T35: Helicobacter (12%); H35: Helicobacter
# T41: Megamonas; H41: Megamonas (41%)

```
